---
title: "Human Plague"
format: html
editor: visual
---

## Exogenous Force of Infection Model

This is a single outbreak model, which forms the general structure of the multiple outbreak, hierarchical latent variable model.

## Packages

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(yersinia)
library(rstan)
library(ggplot2)
library(bayesplot)
rstan_options(auto_write = TRUE)
```

# Data

Uses the Barcelona data set.

```{r}
dat <- outbreaks |>
  dplyr::filter(location == 'Barcelona') |>
  dplyr::select(time = day, deaths)

N_city_dat <- outbreaks |>
  dplyr::filter(location == 'Barcelona') |>
  dplyr::select(population)

N_city <- N_city_dat[[1,1]]

dat_list <- list(
  T      = nrow(dat), 
  t0     = 0.0,
  ts     = as.numeric(dat$time),
  y      = as.integer(dat$deaths),
  N      = as.numeric(N_city),  # scalar
  t_max  = max(dat$time),       # 
  x_r    = numeric(0),          # empty arrays to match Stan
  x_i    = integer(0)
)
```

# Model

This is a much simpler model compared to the Rat-Flea-Human model we have been using. This process models human infections as being driven by an exogenous, time-varying reservoir-to-human hazard (force of infection) $\lambda(t)$ representing the per‑susceptible, per‑day risk of infection arising from the rat–flea system. Here $\lambda(t)$ is parameterized as a Gaussian burst (or pulse) with total area A, peak time t_peak, and width $\sigma$, such that,

$$
\lambda(t) = amp × exp(−(t − t_{peak})^2 / (2 σ^2))
$$

and the instantaneous peak height (amplitude) is derived as,

$$
amp = A/(\sqrt(2\pi)\sigma)
$$

Biologically, $A$ is the cumulative spillover pressure over the outbreak (integrating infectious fleas, bite rates, and per‑bite transmissibility through time), $t_{peak}$ marks when spillover pressure is highest (e.g., near the peak of the rat epidemic), and $\sigma$ controls duration of the hazard (reflecting rat epidemic length, flea survival off‑host, season/climate, and human behavior).

Human infection dynamics follow a simple deterministic system with states $S(t)$ (susceptible), $I(t)$ (infectious), $C_D(t)$ (cumulative deaths), and $C_R(t)$ (cumulative recoveries; which we don't really use). The ODEs are:

$$
\begin{aligned}
\frac{dS}{dt} &= -\lambda h(t)\cdot S(t) \\
\frac{dI}{dt} &= \lambda h(t)\cdot S(t) - \delta \cdot I(t) \\
\frac{dC_D}{dt} &= p_{\text{death}} \cdot \delta \cdot I(t) \\
\frac{dC_R}{dt} &= (1 - p_{\text{death}})\cdot \delta \cdot I(t)
\end{aligned}
$$

where $\delta$ is the infection duration, and $p_{\text{death}}$ is the probability of dying from the infection (these are not identifiable, but have really tight priors on them given what we know about duration of infection and probability of dying from plague infection.

# Stan Code

```{stan output.var="model",data="dat_list",chains=4,iter=1000}
functions {
  // Force of infection function
  real lambda_of_t(real t, real amp, real t_peak, real sigma) {
    real z = (t - t_peak) / sigma;
    return amp * exp(-0.5 * z * z); // Not sure why, but z * z runs faster than z^2
  }
  // ODE system
  array[] real plague_ode(real t, array[] real y, array[] real theta,
                          array[] real x_r, array[] int x_i) {
    real amp     = theta[1]; // amplitude = intensity of spillover, approx num of infectious fleas per human × bite rate × per‑bite transmission probability
    real t_peak  = theta[2]; // peak time = when spillover is at its highest (i.e., time shortly after fleas leave dying rats and seek alternative hosts)
    real sigma   = theta[3]; // hazard width = reflects duration of rat epidemic, flea survival off‑host, climate/seasonality, and human behaviors or interventions
    real delta   = theta[4]; // infection duration
    real p_death = theta[5]; // probability of dying from infection

    real S = y[1];
    real I = y[2];
    real lambda = lambda_of_t(t, amp, t_peak, sigma);

    array[4] real dy;
    dy[1] = -lambda * S;
    dy[2] =  lambda * S - delta * I;
    dy[3] =  p_death * delta * I;
    dy[4] = (1 - p_death) * delta * I;
    return dy;
  }
}
data {
  int<lower=1> T;              // Number of rows
  real t0;
  array[T] real ts;           // time steps
  array[T] int<lower=0> y;    // Deaths
  real<lower=0> N;            // Human population
  real<lower=0> t_max;
  array[0] real x_r;          // empty real data to satisfy signature
  array[0] int x_i;           // empty int data
}
parameters {
  real<lower=0> A;
  real<lower=0, upper=t_max> t_peak;
  real<lower=0> sigma;
  real<lower=0> delta;
  real<lower=0.70, upper=0.98> p_death;
  real<lower=0> phi;
}
transformed parameters {
  array[4] real y0;
  y0[1] = N;  // use scalar N
  y0[2] = 0;
  y0[3] = 0;
  y0[4] = 0;

  real amp = A / (sqrt(2 * pi()) * sigma);
  array[5] real theta = { amp, t_peak, sigma, delta, p_death };

  array[T, 4] real traj = integrate_ode_bdf(plague_ode, y0, t0, ts, theta, x_r, x_i);

  vector[T] mu;
  mu[1] = traj[1, 3];
  for (t in 2:T) mu[t] = traj[t, 3] - traj[t-1, 3];
}
model {
  A       ~ lognormal(log(0.3), 0.5);
  t_peak  ~ uniform(0, t_max);
  sigma   ~ lognormal(log(5), 0.4);
  delta   ~ lognormal(log(0.10), 0.4);
  p_death ~ beta(180, 20); // Very tight prior (70-95% mortality)
  phi     ~ exponential(1);
  y ~ neg_binomial_2(mu + 1e-9, phi);
}
```

# Model Sampling

```{r, message=FALSE, warning=FALSE, results='hide'}
model <- rstan::sampling(model, data = dat_list, cores = 4, chains = 4, iter = 2000, seed = 42)
```

# Model Statistics

Give the summary statistics for the parameters of interest as well as model convergence (e.g., ESS and Rhat) metrics.

```{r}
# Print only selected parameters from a stanfit
print(
  model,
  pars = c("A", "t_peak", "sigma", "delta", "p_death"),
  probs = c(0.05, 0.5, 0.95),
  digits_summary = 2
)
```

# Plots

The first plot shows the posterior density curves.

```{r}
post <- rstan::extract(model)  # permuted = TRUE (default), collapses chains

draws_df <- data.frame(
  A       = post$A,
  t_peak  = post$t_peak,
  sigma   = post$sigma,
  delta   = post$delta,
  p_death = post$p_death
)

dens_long <- tidyr::pivot_longer(draws_df, everything(), names_to = "parameter", values_to = "value")

p_dens <- ggplot(dens_long, aes(x = value)) +
  geom_density(fill = "#2a9d8f", alpha = 0.4, color = "#2a9d8f") +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  labs(title = "Posterior densities", x = NULL, y = "Density") +
  theme_minimal(base_size = 12)

p_dens
```

This second plot shows the relationship between different parameters in the model.

```{r}
post_arr <- rstan::extract(model, permuted = FALSE)  # array: iterations x chains x parameters
param_names <- dimnames(post_arr)[[3]]
idx <- function(name) match(name, param_names)

iter   <- dim(post_arr)[1]
chains <- dim(post_arr)[2]
params <- c("A","t_peak","sigma","delta","p_death")

A_arr       <- post_arr[,, idx("A")]
sigma_arr   <- post_arr[,, idx("sigma")]
t_peak_arr  <- post_arr[,, idx("t_peak")]
delta_arr   <- post_arr[,, idx("delta")]
p_death_arr <- post_arr[,, idx("p_death")]

#amp_arr <- A_arr / (sqrt(2*pi) * sigma_arr)

post_3d <- array(NA_real_, dim = c(iter, chains, length(params)),
                 dimnames = list(NULL, NULL, params))
post_3d[,,1] <- A_arr
post_3d[,,2] <- t_peak_arr
post_3d[,,3] <- sigma_arr
post_3d[,,4] <- delta_arr
post_3d[,,5] <- p_death_arr

bayesplot::color_scheme_set("teal")
p_pairs <- bayesplot::mcmc_pairs(
  post_3d,
  off_diag_args = list(alpha = 0.25, size = 0.4),
  diag_fun = "dens"
)

p_pairs
```

## Posterior predictive check for Malta exogenous model

```{r}
# Extract posterior
post <- rstan::extract(model)  # your fitted Malta model

# Get observed data
obs_deaths <- dat$deaths
n_obs <- length(obs_deaths)

# Generate posterior predictive samples
n_samples <- 500  # number of posterior draws to plot
sample_idx <- sample(1:length(post$A), n_samples)

# Extract trajectories
pred_deaths <- matrix(NA, nrow = n_samples, ncol = n_obs)
for (i in 1:n_samples) {
  idx <- sample_idx[i]
  # Extract mu (predicted deaths) from posterior
  # Note: you'll need to re-extract mu from the stanfit object
  # This assumes you saved mu in generated quantities
  pred_deaths[i, ] <- post$mu[idx, ]
}

# Calculate quantiles
pred_median <- apply(pred_deaths, 2, median)
pred_lower <- apply(pred_deaths, 2, quantile, 0.05)
pred_upper <- apply(pred_deaths, 2, quantile, 0.95)

# Create data frame for plotting
ppc_df <- data.frame(
  time = dat$time,
  observed = obs_deaths,
  median = pred_median,
  lower = pred_lower,
  upper = pred_upper
)

# Plot
ggplot(ppc_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#2a9d8f", alpha = 0.3) +
  geom_line(aes(y = median), color = "#2a9d8f", size = 1) +
  geom_point(aes(y = observed), size = 1.5, alpha = 0.6) +
  labs(
    title = "Posterior Predictive Check: Malta 1813",
    subtitle = "Observed deaths (points) vs. model predictions (line + 90% CI)",
    x = "Day",
    y = "Daily deaths"
  ) +
  theme_minimal(base_size = 12)

# Goodness of fit metrics
residuals <- obs_deaths - pred_median
rmse <- sqrt(mean(residuals^2))
mae <- mean(abs(residuals))

cat("RMSE:", round(rmse, 2), "\n")
cat("MAE:", round(mae, 2), "\n")
```

```{r}
# Posterior predictive check for Malta exogenous model
# Extract parameters from posterior
post <- rstan::extract(model)

# Get mu and phi from posterior
n_samples <- min(500, nrow(post$mu))
sample_idx <- sample(1:nrow(post$mu), n_samples)

# Generate posterior predictive samples manually
pred_deaths <- matrix(NA, nrow = n_samples, ncol = ncol(post$mu))
for (i in 1:n_samples) {
  idx <- sample_idx[i]
  for (t in 1:ncol(post$mu)) {
    pred_deaths[i, t] <- rnbinom(1, 
                                 mu = post$mu[idx, t] + 1e-9, 
                                 size = post$phi[idx])
  }
}

# Calculate quantiles
pred_median <- apply(pred_deaths, 2, median)
pred_lower <- apply(pred_deaths, 2, quantile, 0.05)
pred_upper <- apply(pred_deaths, 2, quantile, 0.95)

# Rest of plotting code...
# Calculate quantiles
pred_median <- apply(pred_deaths, 2, median)
pred_lower <- apply(pred_deaths, 2, quantile, 0.05)
pred_upper <- apply(pred_deaths, 2, quantile, 0.95)

# Create data frame for plotting
ppc_df <- data.frame(
  time = dat$time,
  observed = obs_deaths,
  median = pred_median,
  lower = pred_lower,
  upper = pred_upper
)

# Plot
ggplot(ppc_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#2a9d8f", alpha = 0.3) +
  geom_line(aes(y = median), color = "#2a9d8f", size = 1) +
  geom_point(aes(y = observed), size = 1.5, alpha = 0.6) +
  labs(
    title = "Posterior Predictive Check: Malta 1813",
    subtitle = "Observed deaths vs. posterior predictive distribution (90% interval)",
    x = "Day",
    y = "Daily deaths"
  ) +
  theme_minimal()
```

## Hierarchical Model Prep

```{r}
# Select 4 outbreaks
locations <- c("Malta", 
               #"Prague", 
               "Barcelona")#, 
               #"Florence")

outbreak_data <- outbreaks %>%
  filter(location %in% locations)

# Create outbreak ID mapping
outbreak_mapping <- data.frame(
  location = locations,
  outbreak_num = 1:length(locations)
)

outbreak_data <- outbreak_data %>%
  left_join(outbreak_mapping, by = "location") %>%
  arrange(outbreak_num, day) %>%  # CRITICAL: sort by outbreak_num FIRST, then day
  group_by(outbreak_num) %>%
  mutate(time_rel = day - min(day)) %>%  # Make times start at 0 for each outbreak
  ungroup()

# Verify the sorting is correct
head(outbreak_data, 20)
tail(outbreak_data, 20)

# Get summary info per outbreak
outbreak_info <- outbreak_data %>%
  group_by(outbreak_num, location) %>%
  summarize(
    n_days = n(),
    max_day = max(time_rel),
    population = first(population),
    .groups = "drop"
  ) %>%
  arrange(outbreak_num)

# Compute start/end positions for each outbreak
outbreak_info <- outbreak_info %>%
  mutate(
    end_pos = cumsum(n_days),
    start_pos = lag(end_pos, default = 0) + 1
  )

print(outbreak_info)

# Verify indexing is correct
for (k in 1:nrow(outbreak_info)) {
  cat("Outbreak", k, ":", outbreak_info$location[k], "\n")
  cat("  Start pos:", outbreak_info$start_pos[k], "\n")
  cat("  End pos:", outbreak_info$end_pos[k], "\n")
  cat("  Times:", outbreak_data$time_rel[outbreak_info$start_pos[k]:outbreak_info$end_pos[k]][1:5], "...\n")
}

# Prepare Stan data
dat_hier <- list(
  N_outbreaks = length(locations),
  T_obs = outbreak_info$n_days,
  N_total = nrow(outbreak_data),
  ts = outbreak_data$time_rel,
  y = as.integer(outbreak_data$deaths),
  outbreak_id = outbreak_data$outbreak_num,
  N_pop = outbreak_info$population,
  t_max = outbreak_info$max_day,
  start_pos = outbreak_info$start_pos,
  end_pos = outbreak_info$end_pos,
  x_r = numeric(0),
  x_i = integer(0)
)

str(dat_hier)
```

```{stan output.var="model2",data="dat_list",chains=4,iter=1000}
functions {
  real lambda_of_t(real t, real amp, real t_peak, real sigma) {
    real z = (t - t_peak) / sigma;
    return amp * exp(-0.5 * z * z);
  }
  
  array[] real plague_ode(real t, array[] real y, array[] real theta,
                          array[] real x_r, array[] int x_i) {
    real amp     = theta[1];
    real t_peak  = theta[2];
    real sigma   = theta[3];
    real delta   = theta[4];
    real p_death = theta[5];
    real S = y[1];
    real I = y[2];
    real lambda = lambda_of_t(t, amp, t_peak, sigma);
    array[4] real dy;
    dy[1] = -lambda * S;
    dy[2] =  lambda * S - delta * I;
    dy[3] =  p_death * delta * I;
    dy[4] = (1 - p_death) * delta * I;
    return dy;
  }
}

data {
  int<lower=1> N_outbreaks;
  array[N_outbreaks] int<lower=1> T_obs;
  int<lower=1> N_total;
  array[N_total] real ts;
  array[N_total] int<lower=0> y;
  array[N_total] int<lower=1, upper=N_outbreaks> outbreak_id;
  array[N_outbreaks] real<lower=0> N_pop;
  array[N_outbreaks] real<lower=0> t_max;
  array[N_outbreaks] int<lower=1> start_pos;  // NEW
  array[N_outbreaks] int<lower=1> end_pos;    // NEW
  array[0] real x_r;
  array[0] int x_i;
}

parameters {
  // Hyperparameters
  real mu_log_A;
  real<lower=0> sigma_log_A;
  real mu_t_peak;
  real<lower=0> sigma_t_peak;
  real mu_log_sigma;
  real<lower=0> sigma_log_sigma;
  
  // Outbreak-specific parameters (non-centered)
  vector[N_outbreaks] log_A_raw;
  vector[N_outbreaks] t_peak_raw;
  vector[N_outbreaks] log_sigma_raw;
  
  // Fixed effects (shared across outbreaks)
  real<lower=0> delta;
  real<lower=0.70, upper=0.98> p_death;
  
  // Overdispersion
  real<lower=0> phi;
}

transformed parameters {
  // Non-centered parameterization
  vector[N_outbreaks] log_A = mu_log_A + sigma_log_A * log_A_raw;
  vector[N_outbreaks] A = exp(log_A);
  vector[N_outbreaks] t_peak = mu_t_peak + sigma_t_peak * t_peak_raw;
  vector[N_outbreaks] log_sigma_param = mu_log_sigma + sigma_log_sigma * log_sigma_raw;
  vector[N_outbreaks] sigma_param = exp(log_sigma_param);
  
  // Compute trajectories for each outbreak
  vector[N_total] mu;
  
  for (k in 1:N_outbreaks) {
    array[4] real y0;
    y0[1] = N_pop[k];
    y0[2] = 0;
    y0[3] = 0;
    y0[4] = 0;
    
    real amp = A[k] / (sqrt(2 * pi()) * sigma_param[k]);
    array[5] real theta = { amp, t_peak[k], sigma_param[k], delta, p_death };
    
    // Extract time points for this outbreak using pre-computed indices
    int n_t = T_obs[k];
    int s_pos = start_pos[k];
    int e_pos = end_pos[k];
    array[n_t] real t_k;
    for (i in 1:n_t) {
      t_k[i] = ts[s_pos + i - 1];
    }
    
    // Solve ODE
    array[n_t, 4] real traj = integrate_ode_bdf(plague_ode, y0, -0.01, t_k, theta, x_r, x_i);
    // Extract daily deaths
    mu[s_pos] = traj[1, 3];
    for (i in 2:n_t) {
      mu[s_pos + i - 1] = traj[i, 3] - traj[i-1, 3];
    }
  }
}

model {
  // Hyperpriors (vague)
  mu_log_A ~ normal(log(0.3), 0.5);
  sigma_log_A ~ exponential(1);
  mu_t_peak ~ normal(80, 5);
  sigma_t_peak ~ exponential(0.5);
  mu_log_sigma ~ normal(log(15), 1);
  sigma_log_sigma ~ exponential(1);
  
  // Non-centered parameters
  log_A_raw ~ std_normal();
  t_peak_raw ~ std_normal();
  log_sigma_raw ~ std_normal();
  
  // Fixed effects (tight priors)
  delta ~ lognormal(log(0.10), 0.4);
  p_death ~ uniform(0.70,0.90);
  
  // Overdispersion
  phi ~ exponential(1);
  
  // Likelihood
  y ~ neg_binomial_2(mu + 1e-9, phi);
}

generated quantities {
  // Posterior predictive checks
  array[N_total] int y_rep;
  vector[N_total] log_lik;
  
  for (i in 1:N_total) {
    y_rep[i] = neg_binomial_2_rng(mu[i] + 1e-9, phi);
    log_lik[i] = neg_binomial_2_lpmf(y[i] | mu[i] + 1e-9, phi);
  }
}
```

```{r}
# Fit hierarchical model using the already-compiled model
model_hier <- sampling(
  object = model2,  # Use the compiled stanmodel object
  data = dat_hier,
  chains = 4,
  iter = 1000,
  warmup = 500,
  cores = 4,
  control = list(adapt_delta = 0.95, max_treedepth = 12),
  seed = 42
)
```

# Point Prior 

### No Probability

```{stan output.var=np}
functions {
  // Force of infection function
  real lambda_of_t(real t, real amp, real t_peak, real sigma) {
    real z = (t - t_peak) / sigma;
    return amp * exp(-0.5 * z * z); // Not sure why, but z * z runs faster than z^2
  }
  // ODE system
  array[] real plague_ode(real t, array[] real y, array[] real theta,
                          array[] real x_r, array[] int x_i) {
    real amp     = theta[1]; // amplitude = intensity of spillover, approx num of infectious fleas per human × bite rate × per‑bite transmission probability
    real t_peak  = theta[2]; // peak time = when spillover is at its highest (i.e., time shortly after fleas leave dying rats and seek alternative hosts)
    real sigma   = theta[3]; // hazard width = reflects duration of rat epidemic, flea survival off‑host, climate/seasonality, and human behaviors or interventions
    real delta   = theta[4]; // infection duration
    
    real S = y[1];
    real I = y[2];
    real lambda = lambda_of_t(t, amp, t_peak, sigma);

    array[4] real dy;
    dy[1] = -lambda * S;
    dy[2] =  lambda * S - delta * I;
    dy[3] =  .80 * delta * I;
    dy[4] = (1 - 0.80) * delta * I;
    return dy;
  }
}
data {
  int<lower=1> T;              // Number of rows
  real t0;
  array[T] real ts;           // time steps
  array[T] int<lower=0> y;    // Deaths
  real<lower=0> N;            // Human population
  real<lower=0> t_max;
  array[0] real x_r;          // empty real data to satisfy signature
  array[0] int x_i;           // empty int data
}
parameters {
  real<lower=0> A;
  real<lower=0, upper=t_max> t_peak;
  real<lower=0> sigma;
  real<lower=0> delta;
  real<lower=0> phi;
}
transformed parameters {
  array[4] real y0;
  y0[1] = N;  // use scalar N
  y0[2] = 0;
  y0[3] = 0;
  y0[4] = 0;

  real amp = A / (sqrt(2 * pi()) * sigma);
  array[4] real theta = { amp, t_peak, sigma, delta};

  array[T, 4] real traj = integrate_ode_bdf(plague_ode, y0, t0, ts, theta, x_r, x_i);

  vector[T] mu;
  mu[1] = traj[1, 3];
  for (t in 2:T) mu[t] = traj[t, 3] - traj[t-1, 3];
}
model {
  A       ~ lognormal(log(0.3), 0.5);
  t_peak  ~ uniform(0, t_max);
  sigma   ~ lognormal(log(5), 0.4);
  delta   ~ lognormal(log(0.10), 0.4);
  phi     ~ exponential(1);
  y ~ neg_binomial_2(mu + 1e-9, phi);
}
```

### Barcelona

```{r}
dat <- outbreaks |>
  dplyr::filter(location == 'Barcelona') |>
  dplyr::select(time = day, deaths)

N_city_dat <- outbreaks |>
  dplyr::filter(location == 'Barcelona') |>
  dplyr::select(population)

N_city <- N_city_dat[[1,1]]

dat_list <- list(
  T      = nrow(dat), 
  t0     = 0.0,
  ts     = as.numeric(dat$time),
  y      = as.integer(dat$deaths),
  N      = as.numeric(N_city),  # scalar
  t_max  = max(dat$time),       # 
  x_r    = numeric(0),          # empty arrays to match Stan
  x_i    = integer(0)
)
```

```{r}
model_2 <- rstan::sampling(np, data = dat_list, cores = 4, chains = 4, iter = 2000, seed = 81)
```

```{r}
post_arr <- rstan::extract(model_2, permuted = FALSE)  # array: iterations x chains x parameters
param_names <- dimnames(post_arr)[[3]]
idx <- function(name) match(name, param_names)

iter   <- dim(post_arr)[1]
chains <- dim(post_arr)[2]
params <- c("A","t_peak","sigma","delta")

A_arr       <- post_arr[,, idx("A")]
sigma_arr   <- post_arr[,, idx("sigma")]
t_peak_arr  <- post_arr[,, idx("t_peak")]
delta_arr   <- post_arr[,, idx("delta")]


#amp_arr <- A_arr / (sqrt(2*pi) * sigma_arr)

post_3d <- array(NA_real_, dim = c(iter, chains, length(params)),
                 dimnames = list(NULL, NULL, params))
post_3d[,,1] <- A_arr
post_3d[,,2] <- t_peak_arr
post_3d[,,3] <- sigma_arr
post_3d[,,4] <- delta_arr

bayesplot::color_scheme_set("teal")
p_pairs <- bayesplot::mcmc_pairs(
  post_3d,
  off_diag_args = list(alpha = 0.25, size = 0.4),
  diag_fun = "dens"
)

p_pairs

post_df <- as.data.frame(post_3d[1, , ])

cor(post_df)
cor(post_df)^2
```

### Malta

```{r}
dat <- outbreaks |>
  dplyr::filter(location == 'Malta') |>
  dplyr::select(time = day, deaths)

N_city_dat <- outbreaks |>
  dplyr::filter(location == 'Malta') |>
  dplyr::select(population)

N_city <- N_city_dat[[1,1]]

dat_list <- list(
  T      = nrow(dat), 
  t0     = 0.0,
  ts     = as.numeric(dat$time),
  y      = as.integer(dat$deaths),
  N      = as.numeric(N_city),  # scalar
  t_max  = max(dat$time),       # 
  x_r    = numeric(0),          # empty arrays to match Stan
  x_i    = integer(0)
)
```

```{r}
model_2.M <- rstan::sampling(np, data = dat_list, cores = 4, chains = 4, iter = 2000, seed = 81)
```

```{r}
post_arr <- rstan::extract(model_2.M, permuted = FALSE)  # array: iterations x chains x parameters
param_names <- dimnames(post_arr)[[3]]
idx <- function(name) match(name, param_names)

iter   <- dim(post_arr)[1]
chains <- dim(post_arr)[2]
params <- c("A","t_peak","sigma","delta")

A_arr       <- post_arr[,, idx("A")]
sigma_arr   <- post_arr[,, idx("sigma")]
t_peak_arr  <- post_arr[,, idx("t_peak")]
delta_arr   <- post_arr[,, idx("delta")]


#amp_arr <- A_arr / (sqrt(2*pi) * sigma_arr)

post_3d <- array(NA_real_, dim = c(iter, chains, length(params)),
                 dimnames = list(NULL, NULL, params))
post_3d[,,1] <- A_arr
post_3d[,,2] <- t_peak_arr
post_3d[,,3] <- sigma_arr
post_3d[,,4] <- delta_arr

bayesplot::color_scheme_set("teal")
p_pairs <- bayesplot::mcmc_pairs(
  post_3d,
  off_diag_args = list(alpha = 0.25, size = 0.4),
  diag_fun = "dens"
)

p_pairs

post_df <- as.data.frame(post_3d[1, , ])

cor(post_df)
cor(post_df)^2
```

```{stan output.var=pprior}
functions {
  // Force of infection function
  real lambda_of_t(real t, real amp, real sigma) {
    real z = (t - 80) / sigma;
    return amp * exp(-0.5 * z * z); // Not sure why, but z * z runs faster than z^2
  }
  // ODE system
  array[] real plague_ode(real t, array[] real y, array[] real theta,
                          array[] real x_r, array[] int x_i) {
    real amp     = theta[1]; // amplitude = intensity of spillover, approx num of infectious fleas per human × bite rate × per‑bite transmission probability
    real sigma   = theta[2]; // hazard width = reflects duration of rat epidemic, flea survival off‑host, climate/seasonality, and human behaviors or interventions
    real delta   = theta[3]; // infection duration
    
    real S = y[1];
    real I = y[2];
    real lambda = lambda_of_t(t, amp, sigma);

    array[4] real dy;
    dy[1] = -lambda * S;
    dy[2] =  lambda * S - delta * I;
    dy[3] =  .80 * delta * I;
    dy[4] = (1 - 0.80) * delta * I;
    return dy;
  }
}
data {
  int<lower=1> T;              // Number of rows
  real t0;
  array[T] real ts;           // time steps
  array[T] int<lower=0> y;    // Deaths
  real<lower=0> N;            // Human population
  real<lower=0> t_max;
  array[0] real x_r;          // empty real data to satisfy signature
  array[0] int x_i;           // empty int data
}
parameters {
  real<lower=0> A;
  real<lower=0> sigma;
  real<lower=0> delta;
  real<lower=0> phi;
}
transformed parameters {
  array[4] real y0;
  y0[1] = N;  // use scalar N
  y0[2] = 0;
  y0[3] = 0;
  y0[4] = 0;

  real amp = A / (sqrt(2 * pi()) * sigma);
  array[3] real theta = { amp, sigma, delta};

  array[T, 4] real traj = integrate_ode_bdf(plague_ode, y0, t0, ts, theta, x_r, x_i);

  vector[T] mu;
  mu[1] = traj[1, 3];
  for (t in 2:T) mu[t] = traj[t, 3] - traj[t-1, 3];
}
model {
  A       ~ lognormal(log(0.3), 0.5);
  sigma   ~ lognormal(log(5), 0.4);
  delta   ~ lognormal(log(0.10), 0.4);
  phi     ~ exponential(1);
  y ~ neg_binomial_2(mu + 1e-9, phi);
}
```

```{r}
model_3 <- rstan::sampling(pprior, data = dat_list, cores = 4, chains = 4, iter = 2000, seed = 42)
```

```{r}
post_arr <- rstan::extract(model_3, permuted = FALSE)  # array: iterations x chains x parameters
param_names <- dimnames(post_arr)[[3]]
idx <- function(name) match(name, param_names)

iter   <- dim(post_arr)[1]
chains <- dim(post_arr)[2]
params <- c("A","sigma","delta")

A_arr       <- post_arr[,, idx("A")]
sigma_arr   <- post_arr[,, idx("sigma")]
delta_arr   <- post_arr[,, idx("delta")]


#amp_arr <- A_arr / (sqrt(2*pi) * sigma_arr)

post_3d <- array(NA_real_, dim = c(iter, chains, length(params)),
                 dimnames = list(NULL, NULL, params))
post_3d[,,1] <- A_arr
post_3d[,,2] <- sigma_arr
post_3d[,,3] <- delta_arr

bayesplot::color_scheme_set("teal")
p_pairs <- bayesplot::mcmc_pairs(
  post_3d,
  off_diag_args = list(alpha = 0.25, size = 0.4),
  diag_fun = "dens"
)

p_pairs

post_df <- as.data.frame(post_3d[1, , ])

cor(post_df)
cor(post_df)^2
```
