---
title: "Monty (Barcelona joint fit, 4 outbreaks — Approach B / true hierarchical)"
format: html
editor: visual
---

**Sibling of `monty-barcelona-hierarchical.qmd` (Approach A).** Same data, same model, same R0/alpha/seasonal reparameterization. The difference: **per-outbreak local parameters are drawn from a hyperprior** rather than fitted with independent priors.

For each pooled local parameter `theta` (one of `beta_h, K_r, I_ini, lambda_baseline`):

```
mu_log_theta    ~ Normal(...)    # hyperprior on the log-scale mean
sigma_log_theta ~ Exponential(.) # hyperprior on the log-scale spread
z_theta<g>      ~ Normal(0, 1)   # per-outbreak z-score (NCP)
theta<g>        = exp(mu_log_theta + sigma_log_theta * z_theta<g>)   # derived
```

The non-centered parameterization (z-score form) keeps gradients well-conditioned for HMC and avoids Neal's funnel when `sigma_log_theta` is small. The derived `theta<g>` is computed inside the wrapped packer; it never appears as a fitted parameter (so the prior model is over `mu, sigma, z` only).

What this buys us vs. Approach A:

-   **Partial pooling.** Outbreaks with weak data get pulled toward the cross-outbreak mean. Outbreaks with strong data dominate their own posterior.
-   **Cross-outbreak distribution as output.** The posterior on `(mu_log_theta, sigma_log_theta)` *is* the answer to "how does `theta` vary across centuries in Barcelona?" — a quantity Approach A only gives you implicitly via the spread of independent local posteriors.
-   **Predictive prior for new outbreaks.** A fitted hyperprior gives you a defensible prior for any future outbreak you want to fit, drawn from the same population.

What stays the same:

-   `R0`, `alpha`, `kappa` are still **single global values** (not pooled) — they're properties of plague biology / Barcelona climate / the recording regime, not per-outbreak quantities.
-   `K_h`, `g_h`, `rho` are pinned exactly as in Approach A.
-   The dust2 model (`plague_stochastic_humans`) is unchanged.

Sampling notes:

-   First pass uses `monty_sampler_random_walk` (with adaptive VCV pilot) so we can compare apples-to-apples with Approach A.
-   The right sampler for hierarchical models is **HMC** (`monty_sampler_hmc`) — `monty_dsl` provides analytic gradients for the prior side. Switch when you've confirmed the model structure works. HMC requires unconstrained, differentiable priors throughout — see notes near the prior chunk for the swap-out.

## Setup

```{r}
library(monty)
library(odin2)
library(dust2)

#library(yersinia)
library(tidyverse)
library(lubridate)

library(posterior)
library(bayesplot)

library(sf)
library(stars)
```

## Outbreak selection + paths

Same four Barcelona outbreaks as Approach A. Same data sources.

```{r}
OUTBREAK_IDS <- c(15, 17, 20, 24)

MORTALITY_CSV <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/data-raw/plague-season-v3_2/input/rawdata.csv"
ERA_RDS       <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/era_preprocessed.RDS"
CLIM_REFERENCE_END <- ymd("1990-01-01")
USE_TEMPERATURE   <- TRUE
```

## Data

```{r}
mortality_raw <- read_csv(MORTALITY_CSV, show_col_types = FALSE)

barcelona_raw <- mortality_raw |>
  filter(place == "Barcelona", id %in% OUTBREAK_IDS) |>
  mutate(date = ymd(if_else(str_length(dateorig) < 10,
                            paste0(dateorig, "-01"), dateorig))) |>
  group_by(id) |>
  arrange(date, .by_group = TRUE) |>
  mutate(time = as.integer(date - min(date)) + 1L,
         group = factor(sprintf("Barcelona_%d", id),
                        levels = sprintf("Barcelona_%d", OUTBREAK_IDS))) |>
  ungroup() |>
  select(group, id, year = startyear, time, date, deaths = n, population, lon, lat)

T_max <- max(barcelona_raw$time)

dat <- barcelona_raw |>
  select(group, time, deaths) |>
  complete(group, time = 1:T_max) |>
  arrange(group, time) |>
  mutate(group = as.character(group))

stopifnot(length(unique(table(dat$group))) == 1L,
          all(table(dat$group) == T_max))

cat(sprintf("T_max = %d days; %d groups; %d total rows (NA padded = %d)\n",
            T_max, nlevels(dat$group), nrow(dat), sum(is.na(dat$deaths))))

group_meta <- barcelona_raw |>
  group_by(group) |>
  summarise(id = first(id), year = first(year),
            duration = max(time), n_obs = sum(!is.na(deaths)),
            pop = first(population),
            total_deaths = sum(deaths, na.rm = TRUE),
            peak_deaths  = max(deaths, na.rm = TRUE),
            .groups = "drop")
group_meta
```

## Per-group seasonal vectors

Identical to Approach A. ERA5 climatology → per-outbreak temperature → `delta_R_multiplier`. Each group's seasonal vector is anchored to its own start date.

```{r}
monthly_to_daily_range <- function(monthly, start_date, end_date) {
  stopifnot(length(monthly) == 12)
  start_date <- ymd(start_date); end_date <- ymd(end_date)
  start_year <- year(start_date); end_year <- year(end_date)
  years <- (start_year - 1):(end_year + 1)
  mids <- numeric(0); vals <- numeric(0); offset <- 0
  for (yr in years) {
    yr_start <- ymd(sprintf("%d-01-01", yr))
    yr_end   <- ymd(sprintf("%d-12-31", yr))
    n_days   <- as.integer(yr_end - yr_start) + 1L
    month_starts <- ymd(sprintf("%d-%02d-01", yr, 1:12))
    month_ends   <- c(month_starts[-1] - days(1), yr_end)
    month_mids   <- as.integer(month_starts + (month_ends - month_starts) / 2 - yr_start) + offset
    mids <- c(mids, month_mids)
    vals <- c(vals, monthly)
    offset <- offset + n_days
  }
  fn <- splinefun(mids, vals, method = "periodic")
  full_start <- ymd(sprintf("%d-01-01", years[1]))
  i_start <- as.integer(start_date - full_start) + 1L
  i_end   <- as.integer(end_date   - full_start) + 1L
  fn(i_start:i_end)
}
```

```{r}
groups <- levels(barcelona_raw$group)

if (isTRUE(USE_TEMPERATURE)) {
  barcelona_lonlat <- barcelona_raw |>
    slice(1) |>
    select(lon, lat) |>
    st_as_sf(coords = c("lon", "lat"))

  era <- readRDS(ERA_RDS)

  clim_monthly <- era |>
    st_extract(barcelona_lonlat) |>
    as_tibble() |>
    select(-geometry) |>
    filter(time < CLIM_REFERENCE_END) |>
    mutate(month = month(time, label = TRUE)) |>
    group_by(month) |>
    summarize(t2m = mean(t2m), .groups = "drop") |>
    arrange(month)

  group_seasonal_data <- setNames(
    lapply(groups, function(g) {
      start_date <- min(barcelona_raw$date[barcelona_raw$group == g])
      end_date   <- start_date + days(T_max - 1)
      daily_temp <- monthly_to_daily_range(clim_monthly$t2m, start_date, end_date)
      delta_R_multiplier(daily_temp)
    }),
    groups
  )
} else {
  group_seasonal_data <- setNames(
    lapply(groups, function(g) rep(1, T_max)), groups)
}

## Sanity check (same as Approach A): each per-group seasonal vector should
## be anchored to that outbreak's actual first-observation date.
seasonal_check <- imap_dfr(group_seasonal_data, function(v, g) {
  data_start <- min(barcelona_raw$date[barcelona_raw$group == g])
  expected_day1_temp <- monthly_to_daily_range(
    clim_monthly$t2m, data_start, data_start)
  expected_day1_mult <- delta_R_multiplier(expected_day1_temp)
  tibble(group = g,
         start_date     = data_start,
         start_month    = as.character(month(data_start, label = TRUE)),
         day1_match     = isTRUE(all.equal(v[1], expected_day1_mult)),
         amplitude_ratio = max(v) / min(v))
})
print(seasonal_check)
stopifnot(all(seasonal_check$day1_match),
          all(seasonal_check$amplitude_ratio > 1.3))
```

## Fixed parameters

Same as Approach A: pin `g_h, rho` at literature; pin `K_h` per group from Krauer; everything not fitted falls back to `historical.yaml`.

```{r}
plague <- yersinia::plague_stochastic_humans

## What's fitted as a SCALAR in the packer:
##   - shared_scalar: one value applies to all groups
##       R0, alpha, kappa, plus the hyperparameters (mu_log_*, sigma_log_*)
##   - local_scalar: one value per group (these are the z-scores in NCP form)
##       z_beta_h, z_K_r, z_I_ini, z_lambda_baseline
## All other quantities (g_h, rho, K_h, etc.) are fixed.
pooled_locals <- c("beta_h", "K_r", "I_ini", "lambda_baseline")
local_z_names <- paste0("z_", pooled_locals)

hyper_names <- c(
  paste0("mu_log_",    pooled_locals),
  paste0("sigma_log_", pooled_locals)
)

shared_names <- c("R0", "alpha", "kappa", hyper_names)
fitted_names <- c(shared_names, local_z_names)   # what the SAMPLER explores

## fitted_names_for_strip is what gets stripped from `historical.yaml`. We
## include K_h, beta_r (the wrap will overwrite both), plus the fitted set.
fixed_pars <- plague_fit_fixed_pars(
  scenario = "historical",
  fitted_names = c(pooled_locals, "R0", "alpha", "kappa", "K_h", "beta_r")
)
fixed_pars$beta_I  <- 0
fixed_pars$p_obs   <- 0.8
fixed_pars$R_ini   <- 0
fixed_pars$I_h_ini <- 0
fixed_pars$R_h_ini <- 0
fixed_pars$obs_period <- 1L
fixed_pars$seasonal <- rep(1, T_max)   # overridden per-group by wrap
fixed_pars$g_h     <- 0.1
fixed_pars$rho     <- 2.5

group_fixed_pars <- setNames(
  lapply(groups, function(g) list(K_h = group_meta$pop[group_meta$group == g])),
  groups
)
```

## Hierarchical packer

The packer fits `(R0, alpha, kappa, hyperparams, z_*<g>)` — 3 + 8 + 16 = **27 free parameters**. The wrapped `unpack`:

1. Pulls per-group parameters via `monty_packer_grouped` (so each group gets its hyperparams + z-scores).
2. Computes `theta<g> = exp(mu_log_theta + sigma_log_theta * z_theta<g>)` for each pooled local. These derived values overwrite any same-named entries from the inner unpack.
3. Injects per-group `K_h` (Krauer population).
4. Converts `R0 → beta_r` (R0 reparameterization, same as A).
5. Computes `seasonal[t] = seasonal_data[t]^alpha` per group (continuous seasonality switch, same as A).
6. Strips out the auxiliary names (`R0, alpha, hyperparams, z_*`) so dust2 only sees model parameters.

```{r}
inner_packer <- monty_packer_grouped(
  groups = groups,
  scalar = c(local_z_names, shared_names),
  shared = shared_names,
  fixed  = fixed_pars
)

## IMPORTANT: helpers (R0_to_beta_r, ncp_to_theta) are defined as locals
## inside `with_hierarchical` so they're captured in the closure environment.
## callr workers (monty_runner_callr) serialize the closure with `saveRDS`,
## which carries the enclosing environment but not the global namespace —
## so global helpers would be invisible to worker chains. See CLAUDE.md
## "monty_packer + monty_runner_callr gotcha".
with_hierarchical <- function(packer, group_fixed, group_seasonal,
                              pooled_locals) {
  R0_to_beta_r <- function(R0, g_r, rho, delta_R) {
    R0 * delta_R / ((1 - g_r) * (1 - exp(-rho)))
  }
  ncp_to_theta <- function(mu_log, sigma_log, z) {
    exp(mu_log + sigma_log * z)
  }

  out <- packer
  inner_unpack <- packer$unpack
  out$unpack <- function(x) {
    u <- inner_unpack(x)
    Map(function(g, gp) {
      pars <- c(gp, group_fixed[[g]])

      ## NCP transform: derive each pooled local from its hyperparams + z.
      for (par in pooled_locals) {
        pars[[par]] <- ncp_to_theta(
          pars[[paste0("mu_log_",    par)]],
          pars[[paste0("sigma_log_", par)]],
          pars[[paste0("z_",         par)]]
        )
        pars[[paste0("z_",         par)]] <- NULL
        pars[[paste0("mu_log_",    par)]] <- NULL
        pars[[paste0("sigma_log_", par)]] <- NULL
      }

      ## R0 reparam (same as Approach A).
      pars$beta_r <- R0_to_beta_r(pars$R0, pars$g_r, pars$rho, pars$delta_R)
      pars$R0     <- NULL

      ## Per-group seasonal modulation (same as Approach A).
      pars$seasonal <- group_seasonal[[g]] ^ pars$alpha
      pars$alpha    <- NULL

      pars
    }, names(u), u)
  }
  class(out) <- class(packer)
  out
}

packer <- with_hierarchical(inner_packer, group_fixed_pars,
                            group_seasonal_data, pooled_locals)

cat("packer parameter names (", length(packer$names()), "total):\n")
print(packer$names())
```

## Hierarchical prior

Hyperpriors centred on plausible values from the Approach A pilot:

| Parameter | Hyperprior centre | Reasoning |
|---|---|---|
| `mu_log_beta_h ~ Normal(log(0.03), 1)` | beta_h ≈ 0.03 | Approach A pilot medians 0.02–0.05 |
| `mu_log_K_r ~ Normal(log(15000), 1)` | K_r ≈ 15k | Pilot medians 5k–25k |
| `mu_log_I_ini ~ Normal(log(50), 2)` | I_ini ≈ 50 (wide) | Pilot range 1–500 |
| `mu_log_lambda ~ Normal(log(2), 1)` | lambda ≈ 2 | Pilot medians 1–5 |
| `sigma_log_* ~ Exponential(1)` | E\[sigma\] = 1 | Allows ~2.7x cross-outbreak range at 1 SD; weakly informative |
| `z_*<g> ~ Normal(0, 1)` | standard | NCP convention |

```{r}
shared_prior_exprs <- list(
  bquote(R0    ~ Uniform(2, 12)),
  bquote(alpha ~ Uniform(0, 3)),
  bquote(kappa ~ Gamma(shape = 3, rate = 0.2)),
  ## Hyperpriors — log-scale means and exponential-prior scales.
  bquote(mu_log_beta_h          ~ Normal(.(log(0.03)), 1)),
  bquote(sigma_log_beta_h       ~ Exponential(rate = 1)),
  bquote(mu_log_K_r             ~ Normal(.(log(15000)), 1)),
  bquote(sigma_log_K_r          ~ Exponential(rate = 1)),
  bquote(mu_log_I_ini           ~ Normal(.(log(50)), 2)),
  bquote(sigma_log_I_ini        ~ Exponential(rate = 0.5)),
  bquote(mu_log_lambda_baseline ~ Normal(.(log(2)), 1)),
  bquote(sigma_log_lambda_baseline ~ Exponential(rate = 1))
)

local_z_prior_exprs <- function(g) {
  list(
    bquote(.(as.name(sprintf("z_beta_h<%s>",          g))) ~ Normal(0, 1)),
    bquote(.(as.name(sprintf("z_K_r<%s>",             g))) ~ Normal(0, 1)),
    bquote(.(as.name(sprintf("z_I_ini<%s>",           g))) ~ Normal(0, 1)),
    bquote(.(as.name(sprintf("z_lambda_baseline<%s>", g))) ~ Normal(0, 1))
  )
}

prior_body <- as.call(c(
  list(as.name("{")),
  shared_prior_exprs,
  unlist(lapply(groups, local_z_prior_exprs), recursive = FALSE)
))

prior <- monty_dsl(prior_body)

stopifnot(setequal(prior$parameters, packer$names()))
length(prior$parameters)
```

> **For HMC compatibility:** the priors above use `Uniform` (R0, alpha) and `Gamma` (kappa). HMC needs unconstrained, differentiable priors. To swap in HMC, replace these with:
>
> -   `R0 <- 2 + 10 * inv_logit(R0_logit)` with `R0_logit ~ Normal(0, 1.5)` (or fit on log scale: `log_R0 ~ Normal(log(5), 0.5)`)
> -   `alpha <- 3 * inv_logit(alpha_logit)` with `alpha_logit ~ Normal(0, 1.5)`
> -   `log_kappa ~ Normal(log(15), 0.5)` then `kappa <- exp(log_kappa)`
>
> The pooled hyperpriors here (Normal + Exponential) are already HMC-compatible.

## Initial values

Start every chain at:

-   Hyperparams at their prior means (so the posterior shape doesn't immediately fight the prior).
-   z-scores at 0 (so each `theta<g>` starts at `exp(mu_log_theta)` — the population centre).
-   `R0 = 5, alpha = 1, kappa = 15` (same as Approach A).

```{r}
init_z_centre <- function(g) {
  c(
    setNames(0, sprintf("z_beta_h<%s>",          g)),
    setNames(0, sprintf("z_K_r<%s>",             g)),
    setNames(0, sprintf("z_I_ini<%s>",           g)),
    setNames(0, sprintf("z_lambda_baseline<%s>", g))
  )
}

init_centre <- c(
  R0 = 5, alpha = 1, kappa = 15,
  mu_log_beta_h          = log(0.03), sigma_log_beta_h          = 0.5,
  mu_log_K_r             = log(15000), sigma_log_K_r            = 0.5,
  mu_log_I_ini           = log(50),    sigma_log_I_ini          = 1.0,
  mu_log_lambda_baseline = log(2),     sigma_log_lambda_baseline = 0.5,
  unlist(lapply(groups, init_z_centre))
)
init_centre <- init_centre[packer$names()]

n_chains <- 4
set.seed(1)
init_chains <- replicate(
  n_chains,
  init_centre + rnorm(length(init_centre), 0, 0.05)
)
```

## Diagonal VCV

Step sizes calibrated to each parameter's natural scale. Hyperparameters on log scale → step ~0.1; z-scores → step ~0.3 (since they're Normal(0,1)).

```{r}
hyper_vcv_diag <- c(
  R0 = 0.5^2, alpha = 0.15^2, kappa = 3^2,
  mu_log_beta_h = 0.2^2, sigma_log_beta_h = 0.15^2,
  mu_log_K_r    = 0.2^2, sigma_log_K_r    = 0.15^2,
  mu_log_I_ini  = 0.3^2, sigma_log_I_ini  = 0.2^2,
  mu_log_lambda_baseline = 0.2^2, sigma_log_lambda_baseline = 0.15^2
)
local_z_vcv_diag <- unlist(lapply(groups, function(g) c(
  setNames(0.3^2, sprintf("z_beta_h<%s>",          g)),
  setNames(0.3^2, sprintf("z_K_r<%s>",             g)),
  setNames(0.3^2, sprintf("z_I_ini<%s>",           g)),
  setNames(0.3^2, sprintf("z_lambda_baseline<%s>", g))
)))
vcv_diag <- c(hyper_vcv_diag, local_z_vcv_diag)
vcv_diag <- vcv_diag[packer$names()]
init_vcv <- diag(vcv_diag)
dimnames(init_vcv) <- list(packer$names(), packer$names())
```

## Smoke test

Sanity-check the wired-up packer: at `init_centre`, every group's pars list should have a sensible `beta_h, K_r, I_ini, lambda_baseline, beta_r, seasonal`.

```{r}
unpacked <- packer$unpack(init_centre)

cat("Sample group (Barcelona_15) after unpack:\n")
str(unpacked$Barcelona_15[c("beta_h", "K_r", "I_ini", "lambda_baseline",
                            "beta_r", "K_h")], max.level = 1)
cat("\nseasonal length:", length(unpacked$Barcelona_15$seasonal),
    "; range:", range(unpacked$Barcelona_15$seasonal), "\n")

unfilter <- dust_unfilter_create(plague, time_start = 0,
                                 data = dat, n_groups = length(groups))
ll_init <- dust_likelihood_run(unfilter, unpacked)
cat("\ndeterministic ll per group at init_centre:\n")
print(setNames(ll_init, groups))
cat("sum:", sum(ll_init), "\n")

likelihood_det <- dust_likelihood_monty(unfilter, packer)
posterior_det <- likelihood_det + prior
cat("posterior density at init:", posterior_det$density(init_centre), "\n")
```

## Monty — deterministic pilot (random walk)

Start with `monty_sampler_random_walk` + adaptive VCV — same recipe as Approach A so we can compare. **Switch to `monty_sampler_hmc()` once you confirm the structure works** (and after rewriting the bounded priors to be unconstrained-differentiable, see note above).

```{r}
sampler_det <- monty_sampler_adaptive(initial_vcv = init_vcv)
runner      <- monty_runner_callr(n_workers = n_chains)

samples_pilot <- monty_sample(posterior_det, sampler_det, 50000,
                              initial = init_chains,
                              runner = runner, n_chains = n_chains,
                              burnin = 15000)
```

### Pilot diagnostics

```{r}
samples_pilot_df <- as_draws_df(samples_pilot)
summarise_draws(samples_pilot_df) |> print(n = Inf)

matplot(samples_pilot$density, type = "l", lty = 1,
        xlab = "Sample", ylab = "Log posterior probability density")
```

```{r, fig.width = 10, fig.height = 8}
mcmc_trace(samples_pilot_df, pars = c("R0", "alpha", "kappa"))
```

```{r, fig.width = 10, fig.height = 10}
## Hyperparameter posteriors — the headline output of Approach B.
## mu_log_* tells you the population-level central tendency of each pooled
## local; sigma_log_* tells you how much it varies across outbreaks (small
## sigma = strong shrinkage; big sigma = independent per-outbreak).
mcmc_areas(samples_pilot_df,
           pars = grep("mu_log_|sigma_log_", packer$names(), value = TRUE),
           prob = 0.8, prob_outer = 0.95)
```

```{r, fig.width = 10, fig.height = 12}
## Derived per-group locals (theta<g>) reconstructed from posterior draws.
## These are NOT in samples_pilot_df directly — we have to apply the NCP
## transform to each draw.
derive_thetas <- function(draws_df, pooled_locals, groups) {
  out <- list()
  for (par in pooled_locals) {
    mu_col    <- paste0("mu_log_", par)
    sigma_col <- paste0("sigma_log_", par)
    for (g in groups) {
      z_col      <- sprintf("z_%s<%s>", par, g)
      theta_col  <- sprintf("%s<%s>", par, g)
      out[[theta_col]] <- exp(draws_df[[mu_col]] +
                              draws_df[[sigma_col]] * draws_df[[z_col]])
    }
  }
  bind_cols(draws_df[, c(".chain", ".iteration", ".draw")], as_tibble(out))
}

derived <- derive_thetas(samples_pilot_df, pooled_locals, groups)

derived_long <- derived |>
  pivot_longer(matches("<"),
               names_to = c("param", "group"),
               names_pattern = "(.*)<(.*)>") |>
  mutate(group = factor(group, levels = groups))

ggplot(derived_long, aes(value, group)) +
  ggdist::stat_halfeye(alpha = 0.7) +
  facet_wrap(~ param, scales = "free_x") +
  scale_x_continuous() +
  labs(title = "Pilot — derived per-group locals (after NCP transform)",
       subtitle = "Shrinkage toward the cross-outbreak mean is the hierarchical signal",
       x = NULL, y = NULL) +
  theme_bw()
```

### Pilot fit check (forward simulation)

Same logic as Approach A — push posterior draws through the model and overlay each outbreak's data.

```{r}
n_pp     <- 100
draws_pp <- samples_pilot_df |>
  posterior::thin_draws(thin = max(1, posterior::ndraws(samples_pilot_df) %/% n_pp)) |>
  dplyr::slice_head(n = n_pp)

simulate_one_draw <- function(theta_row, t_grid) {
  unpacked <- packer$unpack(unlist(theta_row[, packer$names()]))
  sim <- vapply(seq_along(unpacked), function(g) {
    pars <- unpacked[[g]]
    sys <- dust_system_create(plague, pars = pars, n_particles = 1)
    dust_system_set_state_initial(sys)
    yy <- dust_system_simulate(sys, t_grid)
    D_h <- drop(dust_unpack_state(sys, yy)$D_h)
    pars$p_obs * D_h + pars$lambda_baseline
  }, numeric(length(t_grid)))
  colnames(sim) <- groups
  sim
}

t_grid <- seq(0, T_max)

deaths_pilot_long <- map_dfr(seq_len(nrow(draws_pp)), function(i) {
  sim <- simulate_one_draw(draws_pp[i, ], t_grid)
  as_tibble(sim) |>
    mutate(time = t_grid, draw = i) |>
    pivot_longer(-c(time, draw), names_to = "group", values_to = "mu") |>
    mutate(group = factor(group, levels = groups))
})

ggplot() +
  geom_line(data = deaths_pilot_long,
            aes(time, mu, group = draw),
            colour = "#22008822") +
  geom_point(data = barcelona_raw,
             aes(time, deaths), size = 0.4) +
  facet_wrap(~ group, scales = "free", ncol = 1) +
  labs(title = "Pilot forward simulation (Approach B)",
       x = "Day", y = "Deaths") +
  theme_bw()
```

### What to look for

After this pilot, check:

1.  **R-hat on hyperparameters.** Mu and sigma should mix cleanly across chains. If sigma_log_* hits zero (chains stuck at the no-variation boundary), you have funnel pathology — switch to HMC.
2.  **Shrinkage.** Compare derived per-group `theta<g>` posteriors (above) with the independent posteriors from Approach A. Outbreaks with weaker data (e.g., short duration, small peaks) should be pulled closer to the cross-outbreak mean.
3.  **Posterior predictive coverage.** The forward simulation should still cover each outbreak's data — partial pooling makes locals more conservative, so very particular outbreaks may show slight tightening of the predictive fan.
4.  **`sigma_log_*` posteriors.** If `sigma_log_X` posterior is concentrated near 0, the data says "this parameter is essentially constant across outbreaks" — informative even if the per-outbreak posteriors look similar.

## Tuned VCV → stochastic production

Identical to Approach A's structure — extract tuned VCV from adaptive sampler, inflate, run stochastic filter, posterior-predictive overlay. (Sections elided here; copy from `monty-barcelona-hierarchical.qmd` if you get this far.)

```{r, eval = FALSE}
det_vcv  <- apply(samples_pilot$details$vcv, c(1, 2), mean)
dimnames(det_vcv) <- list(packer$names(), packer$names())
prod_vcv <- 0.7 * (2.38^2 / length(packer$names())) * det_vcv

filter <- dust_filter_create(plague, time_start = 0,
                             data = dat, n_particles = 1000,
                             n_groups = length(groups), n_threads = 4)
likelihood_final <- dust_likelihood_monty(filter, packer,
                                          save_trajectories = TRUE)
posterior_final  <- likelihood_final + prior
sampler_final    <- monty_sampler_random_walk(prod_vcv)

init_final <- t(as.matrix(samples_pilot_df[
  samples_pilot_df$.iteration == max(samples_pilot_df$.iteration),
  packer$names()]))

samples <- monty_sample(posterior_final, sampler_final, 5000,
                        initial = init_final,
                        runner = runner, n_chains = n_chains, burnin = 500)
```
