---
title: "Monty"
format: html
editor: visual
---

## Setup

```{r}
install.packages(
  c("monty", "dust2", "odin2"),
  repos = c("https://mrc-ide.r-universe.dev", "https://cloud.r-project.org"))

remotes::install_github('flmnh-ai/yersinia')
```

```{r}
pkgbuild::has_build_tools(debug = TRUE)
```

```{r}
library(monty)
library(odin2)
library(dust2)

library(yersinia)
library(tidyverse)

library(posterior)
library(bayesplot)
```

## Outbreak Data

```{r}
dat <- outbreaks |>
  filter(location == 'Barcelona') |>
  select(time = day, deaths)
  
ggplot(dat, aes(time, deaths)) +
  geom_point() +
  theme_bw()

outbreaks |>
  ggplot(aes(day, deaths)) +
  geom_point() +
  facet_wrap(~location, scale = 'free_y') +
  theme_bw()
```

## Odin

Load the stochastic plague model from the package. The odin2 source lives in `inst/odin/plague_stochastic_humans.R` -- this vignette should never inline its own copy, or it will drift from `monty.R` and `run_plague_model()`.

```{r}
plague <- yersinia::plague_stochastic_humans
```

Fixed (non-fitted) parameters come from the historical scenario. `plague_fit_fixed_pars()` is the same helper `monty.R` uses, so the two paths can never drift. Fitted parameters (see `plague_fit_fitted_names()`) are set by the monty packer below and override anything of the same name in `fixed_pars`.

```{r}
fitted_names <- plague_fit_fitted_names()
fitted_names

fixed_pars <- plague_fit_fixed_pars(scenario = "historical",
                                    fitted_names = fitted_names)
fixed_pars
```

## Dust

Run the model through `dust` at the historical fixed parameters (the fitted params use the odin2 defaults from `plague_stochastic_humans.R` here — monty will sweep them later).

```{r}
sys <- dust_system_create(plague, pars = fixed_pars, n_particles = 100)

dust_system_set_state_initial(sys)

t <- seq(0, 200)
y <- dust_system_simulate(sys, t)
deaths <- dust_unpack_state(sys, y)$D_h #|>
 # abind::adrop(1) # for multidimensional arrays

matplot(t, t(deaths), type = "l", lty = 1, col = "#00000033",
        xlab = "Time", ylab = "Deaths")
```

### Particle Filter

Set up a particle filter. The package helper wraps `dust_filter_create()` with the humans generator so you don't have to pass it explicitly.

```{r}
filter <- plague_fit_filter(dat, n_particles = 2000, n_threads = 2)
```

This creates an arbitrary function we can use to generate a likelihood. We need to run this at least once to "activate" the particle filter. Passing `fixed_pars` here matches what the monty packer will plug in for non-fitted params.

```{r}
dust_likelihood_run(filter, fixed_pars)
```

```{r}
get_ll <- function(particles) {
  filter <- plague_fit_filter(dat, n_particles = particles)
  ll <- replicate(100, dust_likelihood_run(filter, c(fixed_pars, list(beta_h = 0.015, beta_r = 0.8, rho = 2.0, g_h = .08, K_r = 4000, K_h = 25000, I_ini = 15))))
  sd(ll)
}

test <- c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600) |>
  map_dbl(get_ll)


plot(log(c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600)), log(test))
plot(c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600), sqrt(test))

```

Look at the filter object.

```{r}
filter
coef(filter)
```

We can use this particle filter as is with the default parameters.

```{r}
huh <- dust_likelihood_run(filter,
                           c(fixed_pars, list(beta_h = 0.04, beta_r = 1.87, rho = .85, g_h = .0437, K_r = 1150, K_h = 15100, I_ini = 20)),
                           save_trajectories = TRUE)
h <- dust_likelihood_last_trajectories(filter)


matplot(dat$time, t(dust_unpack_state(filter, h)$D_h), type = "l",
        lty = 1, col = "#00000044", ylim = c(0, 120),
        xlab = "Time", ylab = "Deaths")

points(deaths ~ time, dat, pch = 19, col = "red")
```

## Monty

Set up our particle filter for use as the likelihood in `monty`. The packer folds `fixed_pars` in via its `fixed` argument so every call into the odin2 model receives the full parameter set.

```{r}
packer <- monty_packer(fitted_names, fixed = fixed_pars)
likelihood <- dust_likelihood_monty(filter, packer, save_trajectories = FALSE)
```

Define the prior. `plague_fit_prior()` returns the monty DSL model used by the SLURM driver; re-defined inline here for visibility.

```{r}
prior <- plague_fit_prior()
prior
```

Combine the likelihood and prior.

```{r}
posterior <- likelihood + prior
```

Diagonal random-walk VCV, same as the one `monty.R` uses.

```{r}
vcv <- plague_fit_vcv()
sampler <- monty_sampler_random_walk(vcv)
```

Run the sampler in parallel with 4 chains. (Equivalent to `plague_fit_setup()` + `monty_sample()` — the shortcut is shown at the end of this section.)

```{r}
runner <- monty_runner_callr(n_workers = 4)
samples <- monty_sample(posterior, sampler, 50000, runner = runner, n_chains = 4, burnin = 2000)
```

> **Shortcut.** The whole filter + packer + prior + sampler assembly is packaged in `plague_fit_setup()`. In a script where you don't need the pedagogical breakdown, the above reduces to:
>
> ``` r
> setup <- plague_fit_setup(dat, scenario = "historical")
> samples <- monty_sample(setup$posterior, setup$sampler, 50000,
>                         runner = monty_runner_callr(n_workers = 4),
>                         n_chains = 4, burnin = 2000)
> ```
>
> That's the exact pattern `monty.R` (the SLURM driver) uses.

### Diagnostics

```{r}
samples_df <- as_draws_df(samples)
summarise_draws(samples_df)
```

Check to see if the chains converged

```{r}
matplot(samples$density, type = "l", lty = 1,
        xlab = "Sample", ylab = "Log posterior probability density")
```

We should probably do some kind of thinning?

```{r}
length(unique(samples$density)) / length(samples$density)
```

```{r}
mcmc_hist(samples_df)
```

```{r}
mcmc_pairs(samples_df)
mcmc_trace(samples_df)
```

We can make a new vcov using the empirical one.

```{r}
param_cols <- samples_df |> 
  select(-c(.chain, .iteration, .draw)) |>  # remove metadata columns
  select(where(is.numeric))  # keep only parameter columns

# Calculate covariance and tune
sample_cov <- cov(param_cols)
d <- ncol(param_cols)
tuned_cov <- (2.38^2 / d) * sample_cov
tuned_cov
```

```{r}
sampler <- monty_sampler_random_walk(tuned_cov)
samples <- monty_sample(posterior, sampler, 10000, runner = runner, n_chains = 4, burnin = 2000)
```

```{r}
summarise_draws(samples_df)
```

```{r}
draws <- as_draws_array(samples)

# Key diagnostic plots:
mcmc_trace(draws)  # Trace plots for all parameters
mcmc_rank_hist(draws)  # Rank histograms (should be flat if mixing well)
#mcmc_rhat_hist(draws)  # Distribution of R-hat values
#mcmc_neff_hist(draws)  # Effective sample sizes

# Pairs plot to see parameter correlations
mcmc_pairs(draws)

# Autocorrelation
mcmc_acf(draws)
```

## Final Reconstruction

```{r}
y <- dust_unpack_state(filter, samples$observations$trajectories)

## Resize the array to remove the chains dimension
deaths <- array(y$D_h, c(nrow(dat), 8000)) # this number needs to be adjusted based on sampler, should find a way to automate!

matplot(dat$time, (deaths), type = "l", lty = 1, col = "#00008822",
        xlab = "Days", ylab = "Deaths")
points(dat$time, dat$deaths, pch = 19, col = "red")
```

```{r}
y$D_h[,,1] |>
  as_tibble() |>
  mutate(time = 1:n()) |>
  pivot_longer(-time) |>
  ggplot(aes(time)) +
  geom_line(aes(y = value, group = name), alpha = 0.01) +
  geom_point(data = dat, aes(y = deaths), color = 'red') +
  theme_bw()
```

Different!

```{r}
mcmc_hist(samples_df)
as_draws_df(samples) |>
  mcmc_hist()
```
