---
title: "Monty (Barcelona joint fit, 4 outbreaks)"
format: html
editor: visual
---

Joint fit of four Barcelona outbreaks (Krauer ids 15, 17, 20, 24) with partial parameter sharing. Three parameters (`R0`, `alpha`, `kappa`) are global — one value used by every outbreak — and three (`beta_h`, `I_ini`, `lambda_baseline`) are local, with one independent value per outbreak. `K_h` and `K_r` are both **pinned per outbreak** to the Krauer-listed population (Didelot-style `K_r = K_h`). `g_h` (human survival probability of infection) and `rho` (carcass infectivity range) are **pinned at literature values** (`g_h = 0.1`, `rho = 2.5`, both from Didelot 2017).

K_r as a free parameter is largely a nuisance for short outbreaks (~200 days, slow rat demography): the rat-side force of infection has K_r cancel in absolute new-infection counts when `T_r ≈ K_r`, and the human-side force has K_r appear only as a `1/K_r` factor that `beta_h` absorbs. Pinning K_r at K_h matches Didelot 2017's choice and makes `beta_h<g>` directly comparable across outbreaks.

**Three reparameterizations vs the earlier `beta_r` joint fit, all done R-side in the wrapped packer (no odin model edits):**

1.  **`R0` instead of `beta_r`.** `R0 = beta_r * (1 - g_r) * (1 - exp(-rho)) / delta_R` is the basic reproduction number; the wrapped packer converts `R0 → beta_r` before pars hit dust2. Better priors (literature R0 ≈ 2–8 for plague), removes the residual `(beta_r, delta_R)` ridge, and the headline output is directly interpretable.
2.  **Per-outbreak temperature-driven seasonal forcing.** Each outbreak gets its own `seasonal_data[t]` vector derived from ERA5 climatology at Barcelona, interpolated across that outbreak's specific calendar window.
3.  **Continuous seasonality switch via `alpha`.** `seasonal_eff[t] = seasonal_data[t]^alpha`, with `alpha` fitted. `alpha = 0` recovers no-forcing behavior; `alpha = 1` matches the literature-derived calibration; `alpha > 1` amplifies seasonality. Replaces the binary `USE_SEASONAL` flag with a continuous parameter whose posterior tells you how strongly the data supports temperature-driven forcing.

This is the "approach A" of the hierarchical fit: cheap parameter sharing via `monty::monty_packer_grouped()`, no random-effects hyperprior. The dust2 grouped filter sums per-group log-likelihoods internally; data is supplied in long format with a `group` column.

> **To revert any of these reparameterizations:** search for `## REPARAM-R0`, `## REPARAM-SEASONAL`, or `## REPARAM-ALPHA` markers in the chunks below. Each is a small, self-contained block that can be commented out independently. Reverting all three brings back the simple `beta_r` + flat-`seasonal` setup of the prior pilot.

## Setup

```{r}
library(monty)
library(odin2)
library(dust2)

#library(yersinia)
library(tidyverse)
library(lubridate)

library(posterior)
library(bayesplot)

## REPARAM-SEASONAL: sf + stars needed for ERA5 climatology extraction.
## To revert: drop these and the climatology section below.
library(sf)
library(stars)
```

## Outbreak selection

Four Barcelona outbreaks from the Krauer dataset, all daily plague-specific and "complete". (Id 16 — the 1475–1476 outbreak — was dropped after a first run; its 421-day double-wave structure pulled the joint fit toward bad regions of parameter space.)

| id  | year | days | population |
|-----|------|------|------------|
| 15  | 1457 | 204  | 40,000     |
| 17  | 1483 | 193  | 30,000     |
| 20  | 1501 | 211  | 25,000     |
| 24  | 1558 | 201  | 35,000     |

```{r}
OUTBREAK_IDS <- c(15, 17, 20, 24)

MORTALITY_CSV <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/data-raw/plague-season-v3_2/input/rawdata.csv"

## REPARAM-SEASONAL: ERA5 climatology RDS for the seasonal forcing vectors.
## To revert (back to flat seasonal = 1): set USE_TEMPERATURE = FALSE.
ERA_RDS <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/era_preprocessed.RDS"
CLIM_REFERENCE_END <- ymd("1990-01-01")  # modern climatology cutoff
USE_TEMPERATURE <- TRUE
```

## Data

Load Krauer raw, filter to the chosen Barcelona ids, then build the long-format frame the dust2 grouped filter expects: columns `group`, `time`, `deaths`. Each outbreak's `time` is days since *its* first observation, 1-indexed (matching the single-outbreak fit convention). Shorter outbreaks are padded with `NA` deaths up to `T_max` so all groups share the same time grid — dust2 requires this and skips the comparison at NA rows.

```{r}
mortality_raw <- read_csv(MORTALITY_CSV, show_col_types = FALSE)

barcelona_raw <- mortality_raw |>
  filter(place == "Barcelona", id %in% OUTBREAK_IDS) |>
  mutate(date = ymd(if_else(str_length(dateorig) < 10,
                            paste0(dateorig, "-01"), dateorig))) |>
  group_by(id) |>
  arrange(date, .by_group = TRUE) |>
  mutate(time = as.integer(date - min(date)) + 1L,
         group = factor(sprintf("Barcelona_%d", id),
                        levels = sprintf("Barcelona_%d", OUTBREAK_IDS))) |>
  ungroup() |>
  ## REPARAM-SEASONAL: keep `date`, `lon`, `lat` for ERA5 extraction below.
  ## To revert: drop date/lon/lat from this select.
  select(group, id, year = startyear, time, date, deaths = n, population, lon, lat)

T_max <- max(barcelona_raw$time)

## dust2 requires the group column to be integer or character (not factor),
## but we want factor ordering for plotting later — keep the factor in
## `barcelona_raw` and coerce to character only in the filter-bound `dat`.
dat <- barcelona_raw |>
  select(group, time, deaths) |>
  complete(group, time = 1:T_max) |>
  arrange(group, time) |>
  mutate(group = as.character(group))

stopifnot(length(unique(table(dat$group))) == 1L,
          all(table(dat$group) == T_max))

cat(sprintf("T_max = %d days; %d groups; %d total rows (NA padded = %d)\n",
            T_max, nlevels(dat$group), nrow(dat), sum(is.na(dat$deaths))))
```

```{r}
ggplot(barcelona_raw, aes(time, deaths)) +
  geom_point(size = 0.5) +
  facet_wrap(~ group, scales = "free", ncol = 1) +
  labs(title = "Barcelona outbreaks (Krauer plague-specific daily)",
       x = "Day since outbreak start", y = "Deaths") +
  theme_bw()
```

Per-outbreak metadata (the `pop` column is wired into the per-group `K_h` below):

```{r}
group_meta <- barcelona_raw |>
  group_by(group) |>
  summarise(id = first(id),
            year = first(year),
            duration = max(time),
            n_obs = sum(!is.na(deaths)),
            pop = first(population),
            total_deaths = sum(deaths, na.rm = TRUE),
            peak_deaths  = max(deaths, na.rm = TRUE),
            .groups = "drop")
group_meta
```

## Per-group seasonal vectors (REPARAM-SEASONAL)

Each outbreak gets its own `seasonal_data[t]` vector — temperature multiplier on `delta_R` derived from ERA5 climatology at Barcelona, interpolated across that outbreak's calendar window. The model computes `delta_R_eff(t) = delta_R * seasonal[t]` (cold weather → multiplier < 1 → carcasses persist longer → higher transmission).

`monthly_to_daily_range()` is copied from `monty-seasonal.qmd` for self-containment (TODO: extract to package). `delta_R_multiplier()` is the package function.

> **To revert (back to flat seasonal = 1):** set `USE_TEMPERATURE <- FALSE` in the setup chunk above. The chunk below produces a list of `rep(1, T_max)` vectors when that flag is off, which is identical to the prior pilot's behavior.

```{r}
## REPARAM-SEASONAL: monthly→daily climatology interpolator (vignette-local).
monthly_to_daily_range <- function(monthly, start_date, end_date) {
  stopifnot(length(monthly) == 12)
  start_date <- ymd(start_date); end_date <- ymd(end_date)
  start_year <- year(start_date); end_year <- year(end_date)
  years <- (start_year - 1):(end_year + 1)            # pad for spline edges
  mids <- numeric(0); vals <- numeric(0); offset <- 0
  for (yr in years) {
    yr_start <- ymd(sprintf("%d-01-01", yr))
    yr_end   <- ymd(sprintf("%d-12-31", yr))
    n_days   <- as.integer(yr_end - yr_start) + 1L
    month_starts <- ymd(sprintf("%d-%02d-01", yr, 1:12))
    month_ends   <- c(month_starts[-1] - days(1), yr_end)
    month_mids   <- as.integer(month_starts + (month_ends - month_starts) / 2 - yr_start) + offset
    mids <- c(mids, month_mids)
    vals <- c(vals, monthly)
    offset <- offset + n_days
  }
  fn <- splinefun(mids, vals, method = "periodic")
  full_start <- ymd(sprintf("%d-01-01", years[1]))
  i_start <- as.integer(start_date - full_start) + 1L
  i_end   <- as.integer(end_date   - full_start) + 1L
  fn(i_start:i_end)
}
```

```{r}
## REPARAM-SEASONAL: ERA5 climatology + per-group temperature multipliers.
if (isTRUE(USE_TEMPERATURE)) {
  barcelona_lonlat <- barcelona_raw |>
    slice(1) |>
    select(lon, lat) |>
    st_as_sf(coords = c("lon", "lat"))

  era <- readRDS(ERA_RDS)

  clim_monthly <- era |>
    st_extract(barcelona_lonlat) |>
    as_tibble() |>
    select(-geometry) |>
    filter(time < CLIM_REFERENCE_END) |>
    mutate(month = month(time, label = TRUE)) |>
    group_by(month) |>
    summarize(t2m = mean(t2m), .groups = "drop") |>
    arrange(month)

  ## Per-group seasonal vector of length T_max. Each group's clock starts on
  ## its own first observation; we extend the temperature multiplier across
  ## T_max days so the dust2 simulation has a value at every step (NA-padded
  ## data positions still need a seasonal value for the model to step).
  group_seasonal_data <- setNames(
    lapply(groups <- levels(barcelona_raw$group), function(g) {
      start_date <- min(barcelona_raw$date[barcelona_raw$group == g])
      end_date   <- start_date + days(T_max - 1)
      daily_temp <- monthly_to_daily_range(clim_monthly$t2m, start_date, end_date)
      delta_R_multiplier(daily_temp)
    }),
    levels(barcelona_raw$group)
  )
} else {
  ## Flat seasonal — equivalent to the prior pilot.
  group_seasonal_data <- setNames(
    lapply(levels(barcelona_raw$group), function(g) rep(1, T_max)),
    levels(barcelona_raw$group)
  )
}

## REPARAM-SEASONAL: per-group sanity check. Each group's seasonal vector
## should start on its own first-observation date and track the calendar
## from there. The model's `seasonal[time + 1]` indexing aligns time = day
## 1 of outbreak with calendar day 1 of that outbreak.
seasonal_check <- imap_dfr(group_seasonal_data, function(v, g) {
  data_start <- min(barcelona_raw$date[barcelona_raw$group == g])
  ## Independently re-derive the day-1 temperature/multiplier from the
  ## climatology — should equal v[1] if the seasonal vector was correctly
  ## anchored to data_start. Catches the failure mode where every group's
  ## seasonal vector accidentally uses the same start date.
  expected_day1_temp <- monthly_to_daily_range(
    clim_monthly$t2m, data_start, data_start)
  expected_day1_mult <- delta_R_multiplier(expected_day1_temp)
  tibble(group = g,
         start_date      = data_start,
         start_month     = as.character(month(data_start, label = TRUE)),
         start_temp_C    = round(expected_day1_temp, 2),
         day1_multiplier = round(v[1], 4),
         expected_mult   = round(expected_day1_mult, 4),
         min_multiplier  = round(min(v), 3),
         max_multiplier  = round(max(v), 3),
         amplitude_ratio = round(max(v) / min(v), 3),
         day1_anchor_ok  = isTRUE(all.equal(v[1], expected_day1_mult)))
})
print(seasonal_check)

## Amplitude ratio < 1.3 would mean weak `alpha` identifiability.
stopifnot(all(seasonal_check$day1_anchor_ok))
stopifnot(all(seasonal_check$amplitude_ratio > 1.3))
cat("\nAll per-group seasonal vectors correctly anchored to outbreak start dates;\n",
    "all seasonal amplitudes > 1.3x (alpha will be identifiable).\n", sep = "")
```

```{r}
## Plot per-group seasonal multiplier on a calendar (month-of-year) axis so
## you can visually verify each outbreak's vector tracks its true season.
seasonal_long <- imap_dfr(group_seasonal_data, function(v, g) {
  start_date <- min(barcelona_raw$date[barcelona_raw$group == g])
  tibble(group = g,
         time  = seq_along(v),
         date  = start_date + days(time - 1),
         month_of_year = month(date) + (mday(date) - 1) / 31,
         seasonal = v)
}) |> mutate(group = factor(group, levels = levels(barcelona_raw$group)))

## Two views: (a) per-group time series with calendar x-axis, (b) all four
## overlaid by month-of-year — confirming they trace the same Barcelona
## climatology, just sampled at different phases.
p_calendar <- ggplot(seasonal_long, aes(date, seasonal)) +
  geom_line() +
  geom_vline(data = seasonal_check, aes(xintercept = start_date),
             linetype = "dashed", colour = "red") +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  labs(title = "Per-outbreak seasonal multiplier (red dash = outbreak start)",
       x = NULL, y = "delta_R multiplier") +
  theme_bw()

p_monthly <- ggplot(seasonal_long, aes(month_of_year, seasonal,
                                       group = group, colour = group)) +
  geom_line(alpha = 0.7) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Seasonal multiplier overlaid by month-of-year",
       subtitle = "All four trace the same Barcelona climatology",
       x = NULL, y = "delta_R multiplier") +
  theme_bw()

print(p_calendar)
print(p_monthly)
```

## Fixed parameters

`historical.yaml` is the right baseline for medieval Barcelona — same era, comparable assumptions. We strip out anything we'll fit (the local + global names below) and override the following:

-   `beta_I = 0` — no human-to-human transmission in this first cut.
-   `R_ini = R_h_ini = I_h_ini = 0` — outbreaks start with all humans susceptible.
-   `p_obs = 0.8` — Didelot convention; partially confounded with `lambda_baseline` so we pin it.
-   `seasonal = rep(1, T_max)` — fallback only; the wrapped packer overrides this per-group with `seasonal_data^alpha` (see Grouped packer section).
-   `g_h = 0.1` and `rho = 2.5` — pinned at Didelot 2017 values (see notebook intro).

```{r}
plague <- yersinia::plague_stochastic_humans

groups <- levels(barcelona_raw$group)

## REPARAM-R0 + REPARAM-ALPHA: shared fitted set is (R0, alpha, kappa)
## instead of (beta_r, kappa). R0 → beta_r conversion and seasonal^alpha
## modulation happen in the wrapped packer below.
## To revert: shared_names <- c("beta_r", "kappa") and remove the R0/alpha
## branches in the with_group_fixed wrap.
shared_names <- c("R0", "alpha", "kappa")
## REPARAM-Kr-PIN: K_r dropped from local_names (was c("beta_h", "K_r",
## "I_ini", "lambda_baseline")) — pinned at K_h per outbreak via
## group_fixed_pars below. To revert: add "K_r" back to local_names and
## remove the K_r entry from group_fixed_pars.
local_names  <- c("beta_h", "I_ini", "lambda_baseline")
fitted_names <- c(shared_names, local_names)

## fitted_names_for_strip is what gets stripped from `historical.yaml`. We
## include `K_h`, `K_r` (both pinned per-group via the wrap) and `beta_r`
## (overwritten from R0 via the wrap).
fixed_pars <- plague_fit_fixed_pars(scenario = "historical",
                                    fitted_names = c(fitted_names, "K_h", "K_r", "beta_r"))
fixed_pars$beta_I  <- 0
fixed_pars$p_obs   <- 0.8
fixed_pars$R_ini   <- 0
fixed_pars$I_h_ini <- 0
fixed_pars$R_h_ini <- 0
fixed_pars$obs_period <- 1L
fixed_pars$seasonal <- rep(1, T_max)   # overridden per-group by wrap
fixed_pars$g_h     <- 0.1
fixed_pars$rho     <- 2.5

## Per-group fixed values: each outbreak gets its own K_h AND K_r pinned
## at the Krauer-listed population (Didelot's K_r = K_h convention). These
## get injected into the per-group pars list by the wrapped packer below.
group_fixed_pars <- setNames(
  lapply(groups, function(g) {
    pop <- group_meta$pop[group_meta$group == g]
    list(K_h = pop, K_r = pop)
  }),
  groups
)
str(group_fixed_pars)
```

## Grouped packer

`monty_packer_grouped()` takes a flat parameter vector — names like `g_h`, `beta_h<Barcelona_15>`, `beta_h<Barcelona_17>`, etc. — and unpacks it into a nested list (one entry per group), each containing the full set of per-group parameters merged with `fixed`. dust2's grouped filter consumes that nested list directly.

The packer's `fixed` argument is global (every group receives the same fixed values), so to pin **per-group** quantities like `K_h` we wrap the packer's `unpack` in a small shim that splices `group_fixed_pars[[g]]` into each group's entry. The wrapped packer keeps the same `names()` (no `K_h<g>` slots in the parameter vector) and dust2 sees the K_h values it needs.

```{r}
inner_packer <- monty_packer_grouped(
  groups = groups,
  scalar = c(local_names, shared_names),
  shared = shared_names,
  fixed  = fixed_pars
)

## Wrap unpack to:
##   (1) inject per-group K_h, K_r (REPARAM-Kr-PIN),
##   (2) convert R0 → beta_r (REPARAM-R0),
##   (3) modulate per-group seasonal_data by alpha (REPARAM-ALPHA + REPARAM-SEASONAL).
## Preserves the monty_packer_grouped class so dust_likelihood_monty accepts it.
##
## IMPORTANT: helpers (R0_to_beta_r) are defined as locals inside this
## function so they're captured in the closure environment. callr workers
## (monty_runner_callr) serialize the closure with `saveRDS`, which carries
## the function's enclosing environment but not the global namespace —
## so a global `R0_to_beta_r` would be invisible to worker chains. See
## CLAUDE.md "monty_packer + monty_runner_callr gotcha".
with_group_fixed <- function(packer, group_fixed, group_seasonal) {
  ## REPARAM-R0: convert R0 (the fitted parameter) to beta_r (what the
  ## model consumes). delta_R, g_r, rho all live in fixed_pars so they're
  ## available per-group. Identity bijection — no information lost.
  R0_to_beta_r <- function(R0, g_r, rho, delta_R) {
    R0 * delta_R / ((1 - g_r) * (1 - exp(-rho)))
  }

  out <- packer
  inner_unpack <- packer$unpack
  out$unpack <- function(x) {
    u <- inner_unpack(x)
    Map(function(g, gp) {
      pars <- c(gp, group_fixed[[g]])
      ## REPARAM-R0
      pars$beta_r <- R0_to_beta_r(pars$R0, pars$g_r, pars$rho, pars$delta_R)
      pars$R0     <- NULL
      ## REPARAM-ALPHA + REPARAM-SEASONAL
      pars$seasonal <- group_seasonal[[g]] ^ pars$alpha
      pars$alpha    <- NULL
      pars
    }, names(u), u)
  }
  class(out) <- class(packer)
  out
}
packer <- with_group_fixed(inner_packer, group_fixed_pars, group_seasonal_data)

cat("packer parameter names (", length(packer$names()), "total):\n")
print(packer$names())

cat("\nsanity check — K_h pinned per group, R0/alpha visible in unpack:\n")
example_theta <- setNames(rep(1, length(packer$names())), packer$names())
example_theta["R0"]    <- 5     # R0 = 5 → beta_r ≈ 1.10 with delta_R=0.2
example_theta["alpha"] <- 1     # full seasonality
example_unpack <- packer$unpack(example_theta)
str(example_unpack$Barcelona_15[c("K_h", "beta_r", "seasonal")], max.level = 1)
```

## Prior

Same prior shapes as the single-outbreak `plague_fit_prior()` for the local parameters, replicated four times. Shared priors:

-   **REPARAM-R0:** `R0 ~ Gamma(shape = 3, rate = 0.6)` — mean 5, mode 3.3, 95% CI ~[1, 11]. Centred on plague R0 literature consensus (Didelot 2017 Cairo posterior ~3–5; Eyam ~1.6–2.2; Black-Death-era estimates 2–3). Smooth and HMC-compatible; gentle pull toward the consensus without hard cutoffs. Replaces the earlier `Uniform(2, 12)`. Reverting: swap back to `Uniform(2, 12)` or to `beta_r ~ Uniform(0.1, 3)` depending on how far you're rolling back.
-   **REPARAM-ALPHA:** `alpha ~ Exponential(rate = 0.5)` — mean 2, no hard upper bound. The previous `Uniform(0, 3)` had alpha railing against the upper bound, so the prior was acting as a cap rather than informing the posterior. Exponential prefers smaller alpha (alpha = 0 recovers no-forcing) but lets the data push as high as it needs. Reverting: swap back to `Uniform(0, 3)` or drop the line entirely.
-   `kappa ~ Gamma(shape = 3, rate = 0.2)` (mean 15) — tightened from `Gamma(2, 0.05)` to match the previous pilot's posterior.

`monty_dsl` won't accept a nested `{}` block, so we build a flat list of `~` statements (one per parameter) and assemble them into a single body expression.

```{r}
local_prior_exprs <- function(g) {
  list(
    bquote(.(as.name(sprintf("beta_h<%s>",          g))) ~ Uniform(0.001, 0.15)),
    ## REPARAM-Kr-PIN: K_r<g> dropped — pinned per group at K_h.
    ## Upper bound widened from 5e3 → 5e4 because the alpha=3 pilot
    ## pushed I_ini against the old 5000 wall (winter-start outbreaks
    ## need huge ignition under strong seasonal suppression).
    bquote(.(as.name(sprintf("I_ini<%s>",           g))) ~ Uniform(1, 5e4)),
    bquote(.(as.name(sprintf("lambda_baseline<%s>", g))) ~ Exponential(rate = 0.1))
  )
}

shared_prior_exprs <- list(
  ## REPARAM-R0: was `Uniform(2, 12)`. Gamma(3, 0.6) has mean 5, mode 3.3,
  ## 95% CI ~[1, 11]. Smooth, HMC-friendly, centred on literature consensus.
  bquote(R0    ~ Gamma(shape = 3, rate = 0.6)),
  ## REPARAM-ALPHA: continuous seasonality switch. Was Uniform(0, 3) —
  ## switched to Exponential because the previous bound was binding.
  ## Mean 2, no hard upper bound; prefers smaller alpha (mode 0 = no
  ## forcing) but lets the data push as needed.
  bquote(alpha ~ Exponential(rate = 0.5)),
  bquote(kappa ~ Gamma(shape = 3, rate = 0.2))
)

prior_body <- as.call(c(
  list(as.name("{")),
  shared_prior_exprs,
  unlist(lapply(groups, local_prior_exprs), recursive = FALSE)
))

prior <- monty_dsl(prior_body)

stopifnot(setequal(prior$parameters, packer$names()))
length(prior$parameters)
```

## Initial values

Per-chain starting values: shared parameters at sensible centres, locals at common centres across outbreaks. Small log-normal jitter across chains keeps R-hat meaningful.

```{r}
init_local_centre <- function(g) {
  c(
    setNames(0.05,  sprintf("beta_h<%s>",          g)),  # near previous posterior
    ## REPARAM-Kr-PIN: K_r dropped from init.
    setNames(1000,  sprintf("I_ini<%s>",           g)),  # bumped from 50 — previous pilot pushed I_ini ~3k–5k
    setNames(3,     sprintf("lambda_baseline<%s>", g))   # near previous posterior median
  )
}

## REPARAM-R0 + REPARAM-ALPHA: starting values bumped to where the previous
## pilot's posterior settled (R0 ~5.2, alpha ~3, kappa ~20). Saves warmup
## time vs starting at the no-seasonal-pilot's hand-fit centres.
init_centre <- c(
  R0 = 5, alpha = 2, kappa = 20,
  unlist(lapply(groups, init_local_centre))
)
init_centre <- init_centre[packer$names()]

n_chains <- 4
set.seed(1)
init_chains <- replicate(
  n_chains,
  init_centre * exp(rnorm(length(init_centre), 0, 0.05))
)
```

## Diagonal VCV

One diagonal VCV entry per parameter, sized in proportion to the prior width. The deterministic adaptive sampler tunes the off-diagonal structure for us, so the diagonal just needs to be in the right ballpark.

```{r}
## REPARAM-R0 + REPARAM-ALPHA: was beta_r = 0.2^2, no alpha entry.
## R0 step ~0.5; alpha step widened from 0.15 to 0.3 (Exponential prior
## now allows higher alpha values — bigger jumps speed exploration).
## REPARAM-Kr-PIN: K_r entry dropped (was 5000^2).
## I_ini step widened from 50^2 to 200^2 to match the wider 1–50000 prior.
vcv_diag <- c(
  R0 = 0.5^2, alpha = 0.3^2, kappa = 3^2,
  unlist(lapply(groups, function(g) c(
    setNames(0.01^2, sprintf("beta_h<%s>",          g)),
    setNames(200^2,  sprintf("I_ini<%s>",           g)),
    setNames(2^2,    sprintf("lambda_baseline<%s>", g))
  )))
)
vcv_diag <- vcv_diag[packer$names()]
init_vcv <- diag(vcv_diag)
dimnames(init_vcv) <- list(packer$names(), packer$names())
```

## Smoke test

Before sampling, run the deterministic likelihood once at `init_centre` to confirm the packer ↔ filter wiring is consistent.

```{r}
unfilter <- dust_unfilter_create(plague, time_start = 0,
                                 data = dat, n_groups = length(groups))
ll_init <- dust_likelihood_run(unfilter, packer$unpack(init_centre))
cat("deterministic ll at init_centre, per group:\n")
print(setNames(ll_init, groups))
cat("sum:", sum(ll_init), "\n")
```

## Monty — deterministic pilot

Same recipe as the single-outbreak fits: deterministic likelihood + adaptive sampler tunes the 15×15 VCV. Four groups means each likelihood call simulates \~4× a single-outbreak call, but the deterministic path is still \~100× cheaper than the stochastic filter.

```{r}
likelihood_det <- dust_likelihood_monty(unfilter, packer)
posterior_det  <- likelihood_det + prior
sampler_det    <- monty_sampler_adaptive(initial_vcv = init_vcv)

runner <- monty_runner_callr(n_workers = n_chains)

samples_pilot <- monty_sample(posterior_det, sampler_det, 30000,
                              initial = init_chains,
                              runner = runner, n_chains = n_chains,
                              burnin = 10000)
```

### Pilot diagnostics

```{r}
samples_pilot_df <- as_draws_df(samples_pilot)
summarise_draws(samples_pilot_df)

matplot(samples_pilot$density, type = "l", lty = 1,
        xlab = "Sample", ylab = "Log posterior probability density")
```

```{r, fig.width = 10, fig.height = 8}
mcmc_trace(samples_pilot_df, pars = shared_names)
mcmc_hist(samples_pilot_df, pars = shared_names)
```

```{r, fig.width = 10, fig.height = 12}
## Per-group local posteriors — one panel per parameter, ridges per outbreak.
samples_pilot_long <- samples_pilot_df |>
  pivot_longer(matches("<"),
               names_to = c("param", "group"),
               names_pattern = "(.*)<(.*)>") |>
  mutate(group = factor(group, levels = groups))

ggplot(samples_pilot_long, aes(value, group)) +
  ## normalize = "panels" rescales each facet's densities to use the full
  ## panel height — without this, facets with large value ranges (K_r,
  ## I_ini) show only the interval line/point because their density
  ## values are squashed by the global rescaling.
  ggdist::stat_halfeye(alpha = 0.7, normalize = "panels") +
  facet_wrap(~ param, scales = "free_x") +
  labs(title = "Pilot — local parameter posteriors by outbreak",
       x = NULL, y = NULL) +
  theme_bw()
```

### Pilot fit check

Forward-simulate from a thinned set of pilot draws and overlay each outbreak's data. If any panel's fan misses its observations, the pilot has settled in the wrong region for that outbreak.

```{r}
n_pp     <- 100
draws_pp <- samples_pilot_df |>
  posterior::thin_draws(thin = max(1, posterior::ndraws(samples_pilot_df) %/% n_pp)) |>
  dplyr::slice_head(n = n_pp)

simulate_one_draw <- function(theta_row, t_grid) {
  unpacked <- packer$unpack(unlist(theta_row[, packer$names()]))
  sim <- vapply(seq_along(unpacked), function(g) {
    pars <- unpacked[[g]]
    sys <- dust_system_create(plague, pars = pars, n_particles = 1)
    dust_system_set_state_initial(sys)
    yy <- dust_system_simulate(sys, t_grid)
    D_h <- drop(dust_unpack_state(sys, yy)$D_h)
    pars$p_obs * D_h + pars$lambda_baseline
  }, numeric(length(t_grid)))
  colnames(sim) <- groups
  sim
}

t_grid <- seq(0, T_max)

deaths_pilot_long <- map_dfr(seq_len(nrow(draws_pp)), function(i) {
  sim <- simulate_one_draw(draws_pp[i, ], t_grid)
  as_tibble(sim) |>
    mutate(time = t_grid, draw = i) |>
    pivot_longer(-c(time, draw), names_to = "group", values_to = "mu") |>
    mutate(group = factor(group, levels = groups))
})

ggplot() +
  geom_line(data = deaths_pilot_long,
            aes(time, mu, group = draw),
            colour = "#22008822") +
  geom_point(data = barcelona_raw,
             aes(time, deaths), size = 0.4) +
  facet_wrap(~ group, scales = "free", ncol = 1) +
  labs(title = "Pilot forward simulation",
       x = "Day", y = "Deaths") +
  theme_bw()
```

### Tuned VCV

```{r}
det_vcv  <- apply(samples_pilot$details$vcv, c(1, 2), mean)
dimnames(det_vcv) <- list(packer$names(), packer$names())

d        <- length(packer$names())
inflate  <- 0.7
prod_vcv <- inflate * (2.38^2 / d) * det_vcv
```

## Monty — production (stochastic)

```{r}
filter <- dust_filter_create(plague, time_start = 0,
                             data = dat, n_particles = 1000,
                             n_groups = length(groups), n_threads = 2)
likelihood_final <- dust_likelihood_monty(filter, packer,
                                          save_trajectories = TRUE)
posterior_final  <- likelihood_final + prior
sampler_final    <- monty_sampler_random_walk(prod_vcv)

init_final <- t(as.matrix(samples_pilot_df[
  samples_pilot_df$.iteration == max(samples_pilot_df$.iteration),
  packer$names()]))

samples <- monty_sample(posterior_final, sampler_final, 5000,
                        initial = init_final,
                        runner = runner, n_chains = n_chains, burnin = 500)

accept_rate <- function(s) {
  apply(s$density, 2, function(x) mean(diff(x) != 0))
}
accept_rate(samples)
```

```{r, fig.width = 10, fig.height = 8}
draws <- as_draws_array(samples)
summarise_draws(draws) |> print(n = Inf)
mcmc_trace(draws, pars = shared_names)
```

## Posterior predictive — per outbreak

```{r}
samples_df <- as_draws_df(samples)

n_pp <- 200
draws_pp <- samples_df |>
  posterior::thin_draws(thin = max(1, posterior::ndraws(samples_df) %/% n_pp)) |>
  dplyr::slice_head(n = n_pp)

deaths_fwd_long <- map_dfr(seq_len(nrow(draws_pp)), function(i) {
  sim <- simulate_one_draw(draws_pp[i, ], t_grid)
  as_tibble(sim) |>
    mutate(time = t_grid, draw = i) |>
    pivot_longer(-c(time, draw), names_to = "group", values_to = "mu") |>
    mutate(group = factor(group, levels = groups))
})

ggplot() +
  geom_line(data = deaths_fwd_long,
            aes(time, mu, group = draw),
            colour = "#88220022") +
  geom_point(data = barcelona_raw,
             aes(time, deaths), size = 0.4) +
  facet_wrap(~ group, scales = "free", ncol = 1) +
  labs(title = "Posterior predictive — forward simulation",
       x = "Day", y = "Deaths") +
  theme_bw()
```

## Cross-outbreak comparison of local parameters

The substantive payoff: do `beta_h`, `I_ini`, `lambda_baseline` differ meaningfully across the four Barcelona centuries, or does the joint posterior pull them toward a common value?

```{r, fig.width = 10, fig.height = 12}
samples_long <- samples_df |>
  pivot_longer(matches("<"),
               names_to = c("param", "group"),
               names_pattern = "(.*)<(.*)>") |>
  mutate(group = factor(group, levels = groups))

ggplot(samples_long, aes(value, group)) +
  ggdist::stat_halfeye(alpha = 0.7, normalize = "panels") +
  facet_wrap(~ param, scales = "free_x") +
  labs(title = "Production — local parameter posteriors by outbreak",
       x = NULL, y = NULL) +
  theme_bw()
```
