---
title: "Monty"
format: html
editor: visual
---

## Setup

```{r}
install.packages(
  c("monty", "dust2", "odin2"),
  repos = c("https://mrc-ide.r-universe.dev", "https://cloud.r-project.org"))

remotes::install_github('flmnh-ai/yersinia')
```

```{r}
pkgbuild::has_build_tools(debug = TRUE)
```

```{r}
library(monty)
library(odin2)
library(dust2)

library(yersinia)
library(tidyverse)

library(posterior)
library(bayesplot)
```

## Outbreak Data

```{r}
dat <- outbreaks |>
  filter(location == 'Barcelona') |>
  select(time = day, deaths)
  
ggplot(dat, aes(time, deaths)) +
  geom_point() +
  theme_bw()
```

## Odin

Rewrite the stochastic plague model for `odin2`.

```{r}
plague <- odin2::odin({
  # Core equations for transitions between compartments:
  update(S) <- S - n_SI + n_susceptible_births - n_deaths_S
  update(I) <- I + n_SI - n_IR - n_deaths_I
  update(R) <- R + n_recovered + n_resistant_births - n_deaths_R
  update(D) <-  n_deaths_S + n_deaths_I + n_deaths_R + (n_IR - n_recovered)
  
  update(F) <- F + n_new_free_fleas - n_fleas_to_rats - n_flea_deaths
  update(N) <- N + flea_growth_rate + n_fleas_to_rats / T_r
  
  # Human compartments
  update(S_h) <- S_h - n_SI_h + births_h - n_deaths_S_h
  update(I_h) <- I_h + n_SI_h - n_IR_h - n_deaths_I_h
  update(R_h) <- R_h + n_recovered_h - n_deaths_R_h
  update(D_h) <-  n_deaths_S_h + n_deaths_I_h + n_deaths_R_h + (n_IR_h - n_recovered_h)
  
  ### Rats
  ## Intermediate calculations
  T_r <- S + I + R  # total rat population
  T_h <- S_h + I_h + R_h  # total human population
  
  
  rat_birth_rate <- r_r * (1 - T_r / K_r)  # per capita
  rat_birth_rate_clipped <- if(rat_birth_rate > 0) rat_birth_rate else 0
  
  ## Flea dynamics
  # Fleas attempt to find rats
  p_flea_to_rat <- 1 - exp(-a * T_r * dt_test)
  n_fleas_to_rats <- Binomial(F, p_flea_to_rat)
  
  # Remaining fleas may die
  remaining_fleas <- F - n_fleas_to_rats
  p_flea_death <- 1 - exp(-d_f * dt_test)
  n_flea_deaths <- Binomial(remaining_fleas, p_flea_death)
  
  ## Infection forces
  lambda_r <- beta_r * n_fleas_to_rats / T_r  # rat infection rate
  # Human infection force based on fleas that don't find rats
  lambda_h <- beta_h * remaining_fleas  # human infection rate
  
  ## Individual probabilities of transition for rats:
  p_SI <- 1 - exp(-lambda_r)  # S to I probability for rats
  p_IR <- 1 - exp(-m_r * dt_test)  # I to R
  p_rat_birth <- 1 - exp(-rat_birth_rate_clipped * dt_test)  # natural rat birth probability
  p_rat_death <- 1 - exp(-d_r * dt_test)  # natural rat death probability
  
  ## Draws from binomial distributions for rats
  n_deaths_S <- Binomial(S, p_rat_death)
  n_deaths_I <- Binomial(I, p_rat_death)
  n_deaths_R <- Binomial(R, p_rat_death)
  n_SI <- Binomial(S - n_deaths_S, p_SI)
  n_IR <- Binomial(I - n_deaths_I, p_IR)  # rats whose disease ends
  n_recovered <- Binomial(n_IR, g_r)  # rats who recover AND survive
  n_births_S <- Binomial(S - n_deaths_S, p_rat_birth)
  n_births_R <- Binomial(R - n_deaths_R, p_rat_birth * 0.75)
  n_resistant_births <- Binomial(n_births_R, p)
  n_susceptible_births <- n_births_S + n_births_R - n_resistant_births
  
  ## Human transitions
  # Human demographic rates
  p_human_death <- 1 - exp(-d_h * dt_test)
  birth_rate_h <- r_h * (1 - T_h / K_h)  # logistic growth
  birth_rate_h_clipped <- if(birth_rate_h > 0) birth_rate_h else 0
  p_human_birth <- 1 - exp(-birth_rate_h_clipped * dt_test)
  
  # Disease transitions
  p_SI_h <- 1 - exp(-lambda_h * dt_test)  # S to I probability for humans
  p_IR_h <- 1 - exp(-m_h * dt_test)
  
  # Actual transitions
  n_deaths_S_h <- Binomial(S_h, p_human_death)
  n_deaths_I_h <- Binomial(I_h, p_human_death)
  n_deaths_R_h <- Binomial(R_h, p_human_death)
  births_h <- Binomial(S_h + R_h, p_human_birth)
  n_SI_h <- Binomial(S_h - n_deaths_S_h, p_SI_h)
  n_IR_h <- Binomial(I_h - n_deaths_I_h, p_IR_h)
  n_recovered_h <- Binomial(n_IR_h, g_h)
  
  ### Fleas
  # dead infected rats x flea index
  n_new_free_fleas <- floor(N * (n_deaths_I + (n_IR - n_recovered)))
  flea_growth_rate <- r_f * N * (1 - N / K_f) * dt_test  # absolute growth rate in flea index
  
  
  deaths <- data()
  deaths ~ Poisson(D_h)
  
  ## Initial states
  initial(S) <- K_r * S_ini
  initial(I) <- I_ini
  initial(R) <- K_r * (1 - S_ini)
  initial(D) <- 0
  initial(N) <- K_f
  initial(F) <- 0
  initial(S_h) <- K_h
  initial(I_h) <- 0
  initial(R_h) <- 0
  initial(D_h) <- 0
  
  ## User defined parameters
  dt_test <- parameter(1/365)
  I_ini <- parameter(10)    # initial infected rats
  S_ini <- parameter(1)     # proportion initial susceptible rats
  K_r <- parameter(2500)    # Rat carrying capacity
  K_h <- parameter(5000)    # Human carrying capacity
  r_r <- parameter(5)       # Rat population growth rate
  r_h <- parameter(0.045)   # Human population growth rate
  p <- parameter(0.975)     # Probability of inherited resistance
  d_r <- parameter(0.2)     # Natural death rate of rats
  d_h <- parameter(0.04)    # Natural death rate of humans
  beta_r <- parameter(4.7)  # Rat infection rate from fleas
  beta_h <- parameter(0.01) # Human infection rate from remaining fleas
  a <- parameter(4e-3)      # Flea search efficiency for rats
  m_r <- parameter(20)      # Infected rat recovery rate
  m_h <- parameter(26)      # Infected human recovery rate
  g_r <- parameter(0.02)    # Probability rat survives infection
  g_h <- parameter(0.1)     # Probability human survives infection
  r_f <- parameter(20)      # Flea reproduction rate
  K_f <- parameter(6.57)    # Flea carrying capacity per rat
  d_f <- parameter(10)      # Death rate of free fleas
})
```

## Dust

This is how we run the model through `dust` with default parameters

```{r}
sys <- dust_system_create(plague, n_particles = 100)

dust_system_set_state_initial(sys)

t <- seq(0, 200)
y <- dust_system_simulate(sys, t)
deaths <- dust_unpack_state(sys, y)$D_h #|>
 # abind::adrop(1) # for multidimensional arrays

matplot(t, t(deaths), type = "l", lty = 1, col = "#00000033",
        xlab = "Time", ylab = "Deaths")
```

### Particle Filter

This is how we setup a particle filter in `dust`

```{r}
filter <- dust_filter_create(plague, time_start = 0, data = dat, n_particles = 2000)
```

This creates an arbitrary function we can use to generate a likelihood. We need to run this at least once to "activate" the particle filter for some reason

```{r}
dust_likelihood_run(filter, list()) # use default parameters with list()
```

```{r}
get_ll <- function(particles) {
  filter <- dust_filter_create(plague, time_start = 0, data = dat, n_particles = particles)
  ll <- replicate(100, dust_likelihood_run(filter, list(beta_h = 0.0075, beta_r = 2, g_h = .15, K_r =1500, K_h = 4000, I_ini = 15)))
  sd(ll)
}

test <- c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600) |>
  map_dbl(get_ll)


plot(log(c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600)), log(test))
plot(c(100, 200, 400, 800, 1600, 3200, 6400, 12800, 25600), sqrt(test))

```

Look at the filter object.

```{r}
filter
```

We can use this particle filter as is with the default parameters.

```{r}
huh <- dust_likelihood_run(filter, list(beta_h = 0.0075, beta_r = 2, g_h = .15, K_r =1500, K_h = 4000, I_ini = 15), save_trajectories = TRUE)
h <- dust_likelihood_last_trajectories(filter)


matplot(dat$time, t(dust_unpack_state(filter, h)$D_h), type = "l",
        lty = 1, col = "#00000044", ylim = c(0, 120),
        xlab = "Time", ylab = "Deaths")

points(deaths ~ time, dat, pch = 19, col = "red")
```

## Monty

Set up our particle filter for use as the likelihood in `monty`.

```{r}
packer <- monty_packer(c("beta_h", "beta_r", "g_h", "K_r", "K_h", "I_ini")) # free parameters
likelihood <- dust_likelihood_monty(filter, packer, save_trajectories = FALSE)
```

Define the prior.

```{r}
prior <- monty_dsl({
  beta_h ~ Uniform(0.001, 0.02)    # around your default ±50%
  beta_r ~ Uniform(1, 6)           # around your default  
  g_h ~ Uniform(0.05, 0.6)         # survival rates
  K_r ~ Uniform(500, 2500)        # rat carrying capacity
  K_h ~ Uniform(1000, 10000)        # human carrying capacity
  I_ini ~ Uniform(1, 50)           # initial infected
})
```

Combine the likelihood and prior.

```{r}
posterior <- likelihood + prior
```

Define a diagonal variance-covariance matrix for the random walk sampler.

```{r}
vcv <- diag(c(0.004^2,   # beta_h
              0.6^2,     # beta_r  
              0.12^2,    # g_h
              500^2,     # K_r
              200^2,     # K_h
              7^2))      # I_ini

sampler <- monty_sampler_random_walk(vcv)
```

Run the sampler in parallel with 4 chains.

```{r}
runner <- monty_runner_callr(n_workers = 6)
samples <- monty_sample(posterior, sampler, 10000, runner = runner, n_chains = 6, burnin = 2000)
```

### Diagnostics

```{r}
samples_df <- as_draws_df(samples)
summarise_draws(samples_df)
```

Check to see if the chains converged

```{r}
matplot(samples$density, type = "l", lty = 1,
        xlab = "Sample", ylab = "Log posterior probability density")
```

We should probably do some kind of thinning?

```{r}
length(unique(samples$density)) / length(samples$density)
```

```{r}
mcmc_hist(samples_df)
```

```{r}
mcmc_pairs(samples_df)
mcmc_trace(samples_df)
```

We can make a new vcov using the empirical one.

```{r}
param_cols <- samples_df |> 
  select(-c(.chain, .iteration, .draw)) |>  # remove metadata columns
  select(where(is.numeric))  # keep only parameter columns

# Calculate covariance and tune
sample_cov <- cov(param_cols)
d <- ncol(param_cols)
tuned_cov <- (2.38^2 / d) * sample_cov
tuned_cov
```

```{r}
sampler <- monty_sampler_random_walk(tuned_cov)
samples <- monty_sample(posterior, sampler, 10000, runner = runner, n_chains = 4, burnin = 2000)
```

```{r}
summarise_draws(samples_df)
```

```{r}
draws <- as_draws_array(samples)

# Key diagnostic plots:
mcmc_trace(draws)  # Trace plots for all parameters
mcmc_rank_hist(draws)  # Rank histograms (should be flat if mixing well)
#mcmc_rhat_hist(draws)  # Distribution of R-hat values
#mcmc_neff_hist(draws)  # Effective sample sizes

# Pairs plot to see parameter correlations
mcmc_pairs(draws)

# Autocorrelation
mcmc_acf(draws)
```

## Final Reconstruction

```{r}
y <- dust_unpack_state(filter, samples$observations$trajectories)

## Resize the array to remove the chains dimension
deaths <- array(y$D_h, c(nrow(dat), 8000)) # this number needs to be adjusted based on sampler, should find a way to automate!

matplot(dat$time, (deaths), type = "l", lty = 1, col = "#00008822",
        xlab = "Days", ylab = "Deaths")
points(dat$time, dat$deaths, pch = 19, col = "red")
```

```{r}
y$D_h[,,1] |>
  as_tibble() |>
  mutate(time = 1:n()) |>
  pivot_longer(-time) |>
  ggplot(aes(time)) +
  geom_line(aes(y = value, group = name), alpha = 0.01) +
  geom_point(data = dat, aes(y = deaths), color = 'red') +
  theme_bw()
```

Different!

```{r}
mcmc_hist(samples_df)
as_draws_df(samples) |>
  mcmc_hist()
```
