---
title: "Monty (Algiers 1817–1819, daily)"
format: html
editor: visual
---

Plague fit with temperature-driven seasonal forcing on carcass decay (`delta_R`), to a **daily** mortality series. Algiers id 9 in the Krauer dataset (Guyon 1855) gives 804 daily plague-death counts from 1817-06-21 to 1819-09-02 — about 2.2 years spanning two epidemic seasons. This is the simpler companion to `monty-london.qmd`: same K&G heritable-resistance machinery (`R_ini > 0`, `r_r > 0`, `d_r > 0`, `g_r` fitted), but on a smaller population (\~100k) and shorter timeline where multi-wave structure is light enough that the deterministic pilot has a fighting chance. Use this notebook to validate the rat-demography parameterisation before pushing it back to the harder London fit.

## Setup

```{r}
install.packages(
  c("monty", "dust2", "odin2"),
  repos = c("https://mrc-ide.r-universe.dev", "https://cloud.r-project.org"))

remotes::install_github('flmnh-ai/yersinia')
```

```{r}
library(monty)
library(odin2)
library(dust2)

#library(yersinia)
library(tidyverse)
library(lubridate)
library(sf)
library(stars)

library(posterior)
library(bayesplot)
```

## Outbreak selection

Algiers 1817-1819 (Krauer id 9). 804 daily plague-mortality counts from Guyon's 1855 transcription. Daily cadence -\> `obs_period = 1L` (the model's default; `D_h` resets every step). Some days are NA (missing record); dust2 skips the comparison at NA observations.

```{r}
OUTBREAK_PLACE <- "Algiers"
OUTBREAK_ID    <- 9

MORTALITY_CSV <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/data-raw/plague-season-v3_2/input/rawdata.csv"
ERA_RDS       <- "~/UF Dropbox/Nick Gauthier/Projects/black-death/data/era_preprocessed.RDS"

# Modern reference window for the ERA5 monthly climatology.
CLIM_REFERENCE_END <- ymd("1990-01-01")

# Seasonal forcing on / off; flip to FALSE to factor it out of the fit.
USE_SEASONAL <- TRUE

# Daily reporting => obs_period = 1 (model default).
OBS_PERIOD <- 1L
```

## Helpers

`monthly_to_daily_range()` interpolates a 12-month climatology to a daily series across an arbitrary date window. `delta_R_multiplier()` turns daily temperature into a multiplier on `delta_R`: 1 at the reference temperature, larger at extremes (faster carcass-infectivity decay). Identical to the helpers in `monty-seasonal.qmd` / `monty-london.qmd`.

```{r}
monthly_to_daily_range <- function(monthly, start_date, end_date) {
  stopifnot(length(monthly) == 12)
  start_date <- ymd(start_date)
  end_date   <- ymd(end_date)
  stopifnot(end_date >= start_date)

  start_year <- year(start_date)
  end_year   <- year(end_date)
  years <- (start_year - 1):(end_year + 1)

  mids   <- numeric(0)
  vals   <- numeric(0)
  offset <- 0

  for (yr in years) {
    feb  <- 28L + leap_year(yr)
    days <- c(31, feb, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)
    mids <- c(mids, cumsum(days) - days / 2 + offset)
    vals <- c(vals, monthly)
    offset <- offset + sum(days)
  }

  daily <- spline(mids, vals, xout = 1:offset)$y

  series_start <- ymd(paste0(start_year - 1, "-01-01"))
  i1 <- as.integer(start_date - series_start) + 1
  i2 <- as.integer(end_date   - series_start) + 1
  daily[i1:i2]
}

delta_R_multiplier <- function(temperature,
                               T_min = 5,
                               T_max = 37,
                               q     = 1.5,
                               T_ref = 17.8,
                               cap   = 1000) {
  if (T_ref <= T_min || T_ref >= T_max) {
    stop("T_ref must lie strictly between T_min and T_max.")
  }
  L_ref <- (T_ref - T_min) * (T_max - T_ref)^q
  in_range <- temperature > T_min & temperature < T_max
  L_T  <- ifelse(in_range,
                 (temperature - T_min) * (T_max - temperature)^q,
                 NA_real_)
  mult <- L_ref / L_T
  pmin(ifelse(is.na(mult), cap, mult), cap)
}
```

## Mortality data

Load the Krauer raw dataset, filter to Algiers id 9, and build the `(time, deaths)` frame the particle filter expects. `time` is in days since the first observation, 1-indexed.

```{r}
mortality_raw <- read_csv(MORTALITY_CSV, show_col_types = FALSE)

outbreak_raw <- mortality_raw |>
  filter(place == OUTBREAK_PLACE, id == OUTBREAK_ID) |>
  mutate(date = ymd(dateorig)) |>
  arrange(date)

stopifnot(nrow(outbreak_raw) > 0,
          unique(outbreak_raw$interval) == "daily")

dat <- outbreak_raw |>
  mutate(time = as.integer(date - min(date)) + 1L) |>
  select(time, deaths = n)

outbreak_lonlat <- outbreak_raw |>
  slice(1) |>
  select(lon, lat) |>
  st_as_sf(coords = c("lon", "lat"))

outbreak_dates <- range(outbreak_raw$date)

cat(sprintf("Outbreak: %s (id %d)\n", OUTBREAK_PLACE, OUTBREAK_ID))
cat(sprintf("Type: %s, complete: %s\n",
            outbreak_raw$type[1], outbreak_raw$complete[1]))
cat(sprintf("Span: %s to %s (%d obs over %d model days; %d NA in deaths)\n",
            outbreak_dates[1], outbreak_dates[2],
            nrow(dat), max(dat$time), sum(is.na(dat$deaths))))
cat(sprintf("Total deaths: %d, peak daily: %d, median daily: %g\n",
            sum(dat$deaths, na.rm = TRUE),
            max(dat$deaths, na.rm = TRUE),
            median(dat$deaths, na.rm = TRUE)))

ggplot(dat |> mutate(date = outbreak_dates[1] + days(time - 1L)),
       aes(date, deaths)) +
  geom_line() +
  labs(title = sprintf("%s outbreak (id %d, daily)", OUTBREAK_PLACE, OUTBREAK_ID),
       x = NULL, y = "Daily plague deaths") +
  theme_bw()
```

## Temperature climatology

Pull ERA5 monthly 2m temperature at Algiers, restrict to the modern reference window, collapse to a 12-month climatology.

```{r}
era <- readRDS(ERA_RDS)

clim_monthly <- era |>
  st_extract(outbreak_lonlat) |>
  as_tibble() |>
  select(-geometry) |>
  filter(time < CLIM_REFERENCE_END) |>
  mutate(month = month(time, label = TRUE)) |>
  group_by(month) |>
  summarize(t2m = mean(t2m), .groups = "drop") |>
  arrange(month)

clim_monthly

ggplot(clim_monthly, aes(month, t2m)) +
  geom_col() +
  labs(title = sprintf("Monthly climatology at %s", OUTBREAK_PLACE),
       x = "Month", y = "t2m (°C)") +
  theme_bw()
```

## Seasonal multiplier

Length must equal `max(dat$time)` so `seasonal[time + 1]` is in-range for every model step.

```{r}
daily_temp <- monthly_to_daily_range(clim_monthly$t2m,
                                     outbreak_dates[1],
                                     outbreak_dates[2])

stopifnot(length(daily_temp) == max(dat$time))

seasonal <- delta_R_multiplier(daily_temp)

if (!USE_SEASONAL) seasonal <- rep(1, max(dat$time))

tibble(time = seq_along(seasonal),
       date = outbreak_dates[1] + days(time - 1L),
       temp = daily_temp,
       seasonal = seasonal) |>
  pivot_longer(c(temp, seasonal)) |>
  ggplot(aes(date, value)) +
  geom_line() +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(title = "Daily temperature and seasonal multiplier",
       x = NULL, y = NULL) +
  theme_bw()
```

## Configuration

Same hybrid as `monty-london.qmd`: Didelot's transmission scales, Keeling & Gilligan's rat demography, with `R_ini`, `r_r`, `d_r`, `g_r` exposed to the sampler. Bounds rescaled for Algiers' smaller population (\~100k vs London 341k; `I_ini` and `R_ini` upper bounds are correspondingly smaller). Algiers is a 2-year, 2-epidemic series so the heritable-resistance mechanism doesn't have to do as much work as in London — useful for verifying the parameterisation produces sensible posteriors before pushing back to the harder fit.

```{r}
plague <- yersinia::plague_stochastic_humans

algiers_population <- unique(outbreak_raw$population)
stopifnot(length(algiers_population) == 1L)

algiers_fitted_names <- c("beta_h", "beta_r", "rho", "delta_R",
                          "beta_I", "I_ini", "kappa", "I_h_ini",
                          "lambda_baseline",
                          "R_ini", "r_r", "d_r", "g_r")

scen <- load_scenario("didelot")
scen$K_r <- algiers_population
scen$K_h <- algiers_population
scen$seasonal <- seasonal
scen$R_h_ini <- 0
scen$obs_period <- OBS_PERIOD

# K&G demography overrides (didelot baseline pins these at zero).
scen$p     <- 0.975
scen$iota  <- 0.75
scen$g_r   <- 0.02
scen$r_r   <- 0.01370
scen$d_r   <- 0.000548

fitted_names <- algiers_fitted_names
fixed_pars <- plague_fit_fixed_pars(scenario = scen,
                                    fitted_names = fitted_names)

stopifnot(identical(fixed_pars$obs_period, OBS_PERIOD),
          length(fixed_pars$seasonal) == max(dat$time))

yersinia:::validate_obs_period(OBS_PERIOD,
                               tau = fixed_pars$tau %||% 1,
                               data_time = dat$time)

str(fixed_pars[c("K_r", "K_h", "m_r", "m_h", "g_h", "p", "iota", "obs_period")])
```

Algiers-scale prior. Same shape as the London prior; `I_ini` and `R_ini` upper bounds reduced to \~5% and \~30% of K_r ≈ 100k respectively.

```{r}
algiers_prior <- monty::monty_dsl({
  beta_h  ~ Uniform(0.001, 0.1)
  beta_r  ~ Uniform(0.1,   3.0)
  rho     ~ Uniform(0.3,   6.0)
  delta_R ~ Uniform(0.05,  1.5)
  beta_I  ~ Uniform(0.0,   0.05)
  ## Initial infected rats. Upper of 5000 = 5% of K_r is the seeding
  ## ceiling; bound below at 10 (a single-rat seed).
  I_ini   ~ Uniform(10, 5000)
  ## NegBinomial overdispersion (size). Tight prior keeps kappa above
  ## ~5; the wider Gamma(2, 0.05) lets it drift to ~0.2 where the
  ## chain absorbs structural misfit as overdispersion noise.
  kappa   ~ Gamma(shape = 10, rate = 0.5)
  I_h_ini ~ Uniform(0,    1000)
  ## Per-day non-plague baseline. With obs_period = 1 the mean
  ## contribution per data point is just `lambda_baseline`. Algiers'
  ## ~100k pop has all-cause mortality of ~8/day, so 1/day misclas-
  ## sification ceiling is generous.
  lambda_baseline ~ Uniform(0, 1)
  ## Initial heritably-resistant rats. Bound at 30% of K_r = 30000.
  R_ini ~ Uniform(0, 30000)
  r_r ~ Uniform(0.005, 0.025)
  d_r ~ Uniform(0.0001, 0.003)
  g_r ~ Uniform(0.001, 0.1)
})

algiers_vcv_diag <- c(
  beta_h  = 0.01^2,
  beta_r  = 0.2^2,
  rho     = 0.5^2,
  delta_R = 0.1^2,
  beta_I  = 0.005^2,
  I_ini   = 500^2,
  kappa   = 5^2,
  I_h_ini = 100^2,
  lambda_baseline = 0.1^2,
  R_ini   = 2000^2,
  r_r     = 0.005^2,
  d_r     = 0.0005^2,
  g_r     = 0.01^2
)

algiers_vcv <- diag(algiers_vcv_diag)
```

## Dust

Sanity-run the model at fixed parameters; with `obs_period = 1`, `D_h` is per-step plague deaths.

```{r}
sys <- dust_system_create(plague, pars = fixed_pars, n_particles = 100)
dust_system_set_state_initial(sys)

t <- seq(0, max(dat$time))
y <- dust_system_simulate(sys, t)
deaths_sim <- dust_unpack_state(sys, y)$D_h

matplot(t, t(deaths_sim), type = "l", lty = 1, col = "#00000033",
        xlab = "Time (days)", ylab = "D_h (daily new plague deaths)")
points(dat$time, dat$deaths, pch = 19, cex = 0.3, col = "red")
```

## Monty — deterministic pilot

```{r}
packer <- monty_packer(fitted_names, fixed = fixed_pars)

unfilter <- dust_unfilter_create(plague, time_start = 0, data = dat)
likelihood_det <- dust_likelihood_monty(unfilter, packer)

prior <- algiers_prior
posterior_det <- likelihood_det + prior

sampler_det <- monty_sampler_adaptive(initial_vcv = algiers_vcv)
```

```{r}
n_chains <- 4

init_full <- c(beta_h  = 0.04, beta_r  = 2.0,  rho     = 2.5,
               delta_R = 0.3,  beta_I  = 0.02, I_ini   = 500,
               kappa   = 20,   I_h_ini = 100,
               lambda_baseline = 0.5,
               R_ini   = 3000,     # ~3% initial heritable resistance
               r_r     = 0.01370,  # K&G default
               d_r     = 0.000548, # K&G default
               g_r     = 0.02)     # K&G default

init_center <- init_full[fitted_names]

set.seed(1)
init_chains <- replicate(n_chains, init_center * exp(rnorm(length(init_center), 0, 0.02)))

runner <- monty_runner_callr(n_workers = n_chains)
samples_pilot <- monty_sample(posterior_det, sampler_det, 40000,
                              initial = init_chains,
                              runner = runner, n_chains = n_chains,
                              burnin = 17000)
```

### Pilot diagnostics

```{r}
samples_pilot_df <- as_draws_df(samples_pilot)
summarise_draws(samples_pilot_df)

matplot(samples_pilot$density, type = "l", lty = 1,
        xlab = "Sample", ylab = "Log posterior probability density")
```

```{r}
#mcmc_pairs(samples_pilot_df)
mcmc_trace(samples_pilot_df)
mcmc_hist(samples_pilot_df)
```

### Pilot fit check

```{r}
n_pp     <- 100
draws_pp <- samples_pilot_df_kept |>
  posterior::thin_draws(thin = max(1, posterior::ndraws(samples_pilot_df_kept) %/% n_pp)) |>
  dplyr::slice_head(n = n_pp)

deaths_pilot <- vapply(seq_len(nrow(draws_pp)), function(i) {
  pars <- modifyList(fixed_pars,
                     as.list(draws_pp[i, fitted_names, drop = FALSE]))
  sys <- dust_system_create(plague, pars = pars, n_particles = 1)
  dust_system_set_state_initial(sys)
  yy <- dust_system_simulate(sys, dat$time)
  D_h <- drop(dust_unpack_state(sys, yy)$D_h)
  pars$p_obs * D_h + pars$lambda_baseline * pars$obs_period
}, numeric(nrow(dat)))

matplot(dat$time, deaths_pilot, type = "l", lty = 1, col = "#22008822",
        xlab = "Days", ylab = "Daily deaths (mu)",
        main = "Forward sim from deterministic pilot (Algiers 1817–1819)")
points(dat$time, dat$deaths, pch = 19, cex = 0.3, col = "black")
```

### Tuned VCV from adaptive pilot

```{r}
det_vcv  <- apply(samples_pilot$details$vcv[, , keep_chain, drop = FALSE], c(1, 2), mean)
dimnames(det_vcv) <- list(fitted_names, fitted_names)

d        <- length(fitted_names)
inflate  <- 0.7
prod_vcv <- inflate * (2.38^2 / d) * det_vcv
```

## Monty — production (stochastic)

```{r}
filter <- plague_fit_filter(dat, n_particles = 500, n_threads = 2)
likelihood_final <- dust_likelihood_monty(filter, packer,
                                          save_trajectories = TRUE)
posterior_final <- likelihood_final + prior

sampler_final <- monty_sampler_random_walk(prod_vcv)

last_iter <- max(samples_pilot_df_kept$.iteration)
final_pool <- samples_pilot_df_kept |>
  dplyr::filter(.data$.iteration == last_iter)
if (nrow(final_pool) < n_chains) {
  pool_late <- samples_pilot_df_kept |>
    dplyr::filter(.data$.iteration > last_iter - 200)
  set.seed(2)
  filler <- pool_late[sample(nrow(pool_late), n_chains - nrow(final_pool)), ]
  final_pool <- dplyr::bind_rows(final_pool, filler)
}
init_final <- t(as.matrix(final_pool[, fitted_names]))

samples <- monty_sample(posterior_final, sampler_final, 5000,
                        initial = init_final,
                        runner = runner, n_chains = n_chains, burnin = 500)

accept_rate <- function(s) {
  d <- s$density
  apply(d, 2, function(x) mean(diff(x) != 0))
}

accept_rate(samples)

draws <- as_draws_array(samples)
summarise_draws(draws)
mcmc_trace(draws)
mcmc_hist(draws)
```

## Posterior predictive: forward simulation

```{r}
samples_df <- as_draws_df(samples)

n_pp <- 200
draws_pp <- samples_df |>
  posterior::thin_draws(thin = max(1, posterior::ndraws(samples_df) %/% n_pp)) |>
  dplyr::slice_head(n = n_pp)

deaths_fwd <- vapply(seq_len(nrow(draws_pp)), function(i) {
  pars <- modifyList(fixed_pars,
                     as.list(draws_pp[i, fitted_names, drop = FALSE]))
  sys <- dust_system_create(plague, pars = pars, n_particles = 1)
  dust_system_set_state_initial(sys)
  yy <- dust_system_simulate(sys, dat$time)
  D_h <- drop(dust_unpack_state(sys, yy)$D_h)
  pars$p_obs * D_h + pars$lambda_baseline * pars$obs_period
}, numeric(nrow(dat)))

matplot(dat$time, deaths_fwd, type = "l", lty = 1, col = "#88220022",
        xlab = "Days", ylab = "Daily deaths (mu)",
        main = "Forward simulation from posterior draws")
points(dat$time, dat$deaths, pch = 19, cex = 0.3, col = "black")
```

## Filter-conditioned reconstruction

```{r}
dust_likelihood_run(filter, fixed_pars)

y <- dust_unpack_state(filter, samples$observations$trajectories)
deaths_filt <- array(y$D_h, c(nrow(dat), prod(dim(y$D_h)[-1])))
matplot(dat$time, deaths_filt, type = "l", lty = 1, col = "#00008822",
        xlab = "Days", ylab = "Daily D_h",
        main = "Filter-conditioned trajectories")
points(dat$time, dat$deaths, pch = 19, cex = 0.3, col = "red")
```
