---
title: "Introduction to yersinia: Plague Transmission Modeling in R"
author: "Nicolas Gauthier"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Introduction to yersinia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.retina = 2
)
```

# Overview

The `yersinia` package provides a comprehensive toolkit for modeling plague transmission dynamics using stochastic simulation. Built on the `odin2` / `dust2` / `monty` stack, it uses a carcass-based transmission formulation (Didelot et al. 2017) to capture realistic biological processes including demographic stochasticity and multi-host dynamics.

## Why Stochastic Models?

Plague outbreaks exhibit high variability due to:
- Small population sizes where random events matter
- Environmental stochasticity in transmission rates
- Stochastic extinction and recolonization events

Traditional deterministic models miss these crucial dynamics that determine real outbreak patterns.

## Key Features

- **Stochastic simulation**: Demographic noise and realistic extinction/recolonization dynamics
- **Evidence-based parameters**: Curated parameter sets from historical and contemporary research
- **Multi-host dynamics**: Carcass-based transmission with optional human dynamics
- **Bayesian fitting**: Particle-filter pMCMC via `monty` for fitting to historical death-count time series (see `vignette("monty")`)
- **Comprehensive analysis**: Built-in functions for R₀, outbreak metrics, force of infection, and visualization
- **Professional plotting**: Phase portraits, comparison plots, and publication-ready figures

```{r setup}
#library(yersinia)
library(dplyr)
library(ggplot2)
library(purrr)

# Set a random seed for reproducible results
set.seed(42)
```

# Getting Started

## Your First Plague Model

The simplest way to run a plague model is with `run_plague_model()`:

```{r basic-model}
# Run a stochastic plague model (single population)
results <- run_plague_model(
  years = 2,
  n_particles = 100           # Stochastic replicates for uncertainty quantification
)

# View the structured results
results
```

The model returns a `plague_results` object with:
- **Tidy data format**: Time-series data in long format with standardized column names
- **Metadata**: Model type, parameters, and run information stored as attributes
- **Multiple replicates**: Each particle represents one stochastic realization

## Built-in Plotting

All results have intelligent default plotting that shows uncertainty across replicates:

```{r basic-plot, fig.cap="Basic plague dynamics showing median trajectories with 95% confidence intervals"}
plot(results)
```

## Extracting Key Information

The package provides professional analysis tools instead of manual calculations:

```{r basic-analysis}
# Get outbreak summary statistics
outbreak_stats <- calculate_outbreak_metrics(results, compartment = "I")
summary_stats <- summarize_outbreak_metrics(outbreak_stats)

print(summary_stats)
```

# Evidence-Based Parameters

## Curated Parameter Sets

The package includes parameter sets from key plague research:

```{r parameter-sets}
# Examine a parameter set
keeling_params <- load_scenario("keeling-gilligan")
keeling_params

# Compare R₀ across all parameter sets
param_sets <- c("defaults", "keeling-gilligan", "modern-estimates", "historical")

r0_comparison <- tibble(
  Parameter_Set = param_sets,
  R0 = map_dbl(param_sets, ~ calculate_R0(load_scenario(.x))),
  Source = c("Package defaults", "Keeling & Gilligan (2000)", "Contemporary estimates", "Medieval estimates")
)

print(r0_comparison)
```

The basic reproduction number (R₀) determines outbreak potential:
- **R₀ > 1**: Disease can spread and establish
- **R₀ < 1**: Disease will die out without sustained transmission

## Parameter Customization

Parameters can be modified at runtime without editing scenario files:

```{r custom-parameters}
# Run model with modified transmission parameters
results_enhanced <- run_plague_model(
  scenario = "keeling-gilligan",
  beta_r = 0.05,         # Higher rat infection rate (per day)
  g_r = 0.05,            # Higher rat survival rate
  rho = 12.0,            # Wider carcass infectivity range
  years = 3,
  n_particles = 80
)

# Compare outbreak metrics
enhanced_stats <- results_enhanced |>
  calculate_outbreak_metrics(compartment = "I") |>
  summarize_outbreak_metrics()

cat("Enhanced transmission outbreak probability:", enhanced_stats$outbreak_probability, "\n")
```

# Model Architecture

## Single Population: Core Dynamics

Examine the fundamental carcass-based transmission cycle:

```{r single-population}
# Focused plague dynamics
results_core <- run_plague_model(
  scenario = "keeling-gilligan",
  n_particles = 60,
  years = 5
)

# Show only core compartments for clarity
plot_dynamics(results_core, compartments = c("S", "I", "R"))
```

### Phase Space Analysis

Visualize the dynamical system in phase space:

```{r phase-portrait, fig.cap="Phase portrait showing susceptible-infected dynamics"}
# Create phase portrait for S-I dynamics
# Note: Advanced plotting functions are provisional and may change in future versions
plot_phase_portrait(results_core, compartments = c("S", "I"))
```

## Multi-Host Dynamics: Rats + Humans

Include human populations in the transmission cycle:

```{r human-model, fig.cap="Multi-host plague dynamics with humans"}
# Model including human transmission
results_humans <- run_plague_model(
  scenario = "historical",     # Medieval parameters for dramatic effect
  include_humans = TRUE,
  n_particles = 60,
  years = 8
)

# Plot all compartments to show multi-host dynamics
plot_dynamics(results_humans)
```

### Human-Specific Analysis

Focus on human epidemic patterns:

```{r human-analysis}
# Calculate human outbreak metrics
human_metrics <- results_humans |>
  calculate_outbreak_metrics(compartment = "Ih") |>
  summarize_outbreak_metrics()

cat("Human outbreak probability:", round(human_metrics$outbreak_probability, 3), "\n")
cat("Mean peak human infections:", round(human_metrics$mean_peak, 0), "\n")
```

## Scenario Comparison

Compare dynamics across different evidence-based parameter sets:

```{r scenario-comparison, fig.cap="Comparison of plague dynamics across parameter scenarios"}
# Compare modern vs historical parameter estimates
results_modern <- run_plague_model(
  scenario = "modern-estimates",
  years = 12,
  n_particles = 50
)

results_historical_long <- run_plague_model(
  scenario = "historical",
  years = 12,
  n_particles = 50
)

# Plot comparison using built-in function
plot_comparison(list("Modern" = results_modern, "Historical" = results_historical_long),
                compartment = "I")
```

# Advanced Analysis Toolkit

## Reproduction Numbers and Invasion Analysis

Calculate R₀ to predict outbreak potential:

```{r r0-analysis}
# R₀ comparison across scenarios
r0_analysis <- tibble(
  Scenario = c("Historical", "Modern", "High Transmission", "Slow Decay"),
  Parameters = list(
    load_scenario("historical"),
    load_scenario("modern-estimates"),
    load_scenario("modern-estimates", beta_r = 1.5),
    load_scenario("modern-estimates", delta_R = 0.1)
  ),
  R0 = map_dbl(Parameters, calculate_R0),
  Invasion_Risk = ifelse(R0 > 1, "High", "Low")
)

print(r0_analysis[c("Scenario", "R0", "Invasion_Risk")])

# Visualize invasion thresholds
ggplot(r0_analysis, aes(x = reorder(Scenario, R0), y = R0, fill = Invasion_Risk)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", size = 1) +
  coord_flip() +
  scale_fill_manual(values = c("High" = "coral", "Low" = "lightblue")) +
  labs(title = "Outbreak Risk Assessment",
       subtitle = "R₀ > 1 indicates potential for sustained transmission",
       x = "Scenario", y = "Basic Reproduction Number (R₀)") +
  theme_minimal()
```

## Force of Infection Dynamics

Analyze transmission pressure over time:

```{r force-infection, eval=FALSE, fig.cap="Force of infection dynamics showing transmission pressure"}
# Reconstruct per-step force of infection from a multi-host run
foi_data <- calculate_force_of_infection(results_humans)

foi_data |>
  tidyr::pivot_longer(c(lambda_h, lambda_r), names_to = "host", values_to = "foi") |>
  ggplot(aes(time, foi, group = replicate)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~host, scales = "free_y",
             labeller = labeller(host = c("lambda_h" = "Humans", "lambda_r" = "Rats"))) +
  labs(title = "Force of Infection Over Time",
       x = "Time (years)", y = "Force of Infection") +
  theme_minimal()
```

## Professional Outbreak Metrics

Use built-in analysis functions for comprehensive outbreak characterization:

```{r outbreak-metrics}
# Comprehensive outbreak analysis for the rats+humans model
outbreak_metrics <- results_humans |>
  calculate_outbreak_metrics(compartment = "I", threshold = 5) |>
  summarize_outbreak_metrics()

# Display key metrics
cat("Rats+Humans Outbreak Summary:\n")
cat("=============================\n")
cat("Outbreak probability:", round(outbreak_metrics$outbreak_probability[1], 3), "\n")
cat("Mean peak infections:", round(outbreak_metrics$mean_peak[1], 1), "\n")
cat("Mean outbreak duration:", round(outbreak_metrics$mean_duration[1], 1), "years\n")
cat("Time to peak:", round(outbreak_metrics$mean_time_to_peak[1], 1), "years\n")
```

# Parameter Sensitivity & Model Comparison

## Systematic Parameter Exploration

Examine how key parameters influence outbreak dynamics:

```{r parameter-sensitivity, fig.cap="Parameter sensitivity analysis across key transmission parameters"}
# Systematic sensitivity analysis
sensitivity_params <- tidyr::expand_grid(
  beta_r = c(0.3, 0.5, 0.77, 1.0, 1.5),
  rho = c(1.0, 2.0, 2.63, 4.0, 6.0)
) |>
  mutate(scenario_id = paste0("beta", beta_r, "_rho", rho))

# Run models and calculate outbreak metrics
sensitivity_analysis <- sensitivity_params |>
  mutate(
    results = pmap(list(beta_r, rho), ~ run_plague_model(
      scenario = "defaults",
      beta_r = ..1,
      rho = ..2,
      n_particles = 30,
      years = 6
    )),
    metrics = map(results, ~ calculate_outbreak_metrics(.x, compartment = "I")),
    summary = map(metrics, summarize_outbreak_metrics),
    outbreak_prob = map_dbl(summary, ~ .x$outbreak_probability[1]),
    mean_peak = map_dbl(summary, ~ .x$mean_peak[1])
  )

# Visualize parameter sensitivity
ggplot(sensitivity_analysis, aes(beta_r, rho, fill = outbreak_prob)) +
  geom_tile() +
  geom_text(aes(label = round(outbreak_prob, 2)), color = "white", size = 3) +
  scale_fill_viridis_c(name = "Outbreak\nProbability") +
  labs(title = "Parameter Sensitivity: Outbreak Probability",
       x = "Carcass-to-Rat Transmission Rate (beta_r, per day)",
       y = "Carcass Infectivity Range (rho)") +
  theme_minimal()
```

## Model Architecture Comparison  

Compare dynamics across different model structures:

```{r architecture-comparison, fig.cap="Comparison of model architectures"}
# Define different model configurations
model_configs <- list(
  "Rats only"            = list(include_humans = FALSE),
  "Rats + Humans"        = list(include_humans = TRUE),
  "Historical (rats)"    = list(scenario = "historical", include_humans = FALSE),
  "Historical (humans)"  = list(scenario = "historical", include_humans = TRUE)
)

# Run all model configurations
model_results <- imap(model_configs, ~ {
  args <- .x
  if (!"scenario" %in% names(args)) args$scenario <- "modern-estimates"
  args$n_particles <- 40
  args$years <- 8
  do.call(run_plague_model, args)
})

# Use built-in comparison plotting
plot_comparison(model_results, compartment = "I")
```

## Equilibrium Analysis

Examine long-term system behavior:

```{r equilibrium-analysis}
# Calculate theoretical equilibria
# equilibrium_comparison <- param_sets |>
#   map(~ {
#     params <- load_scenario(.x)
#     eq <- calculate_equilibrium(params, model_type = "rats_only")
#     tibble(
#       parameter_set = .x,
#       R0 = eq$R0,
#       equilibrium_type = eq$equilibrium_type,
#       endemic_infected = ifelse(eq$equilibrium_type == "endemic", eq$I_r, 0)
#     )
#   }) |>
#   bind_rows()
# 
# print(equilibrium_comparison)
```

# Historical Applications

## Black Death Modeling

Model medieval plague outbreaks with period-appropriate parameters:

```{r black-death-scenario, fig.cap="Black Death scenario with multi-host dynamics"}
# Historical plague with human transmission
results_black_death <- run_plague_model(
  scenario = "historical",
  include_humans = TRUE,
  n_particles = 60,
  years = 6
)

# Analyze human epidemic characteristics
human_outbreak <- results_black_death |>
  calculate_outbreak_metrics(compartment = "Ih") |>
  summarize_outbreak_metrics()

cat("Black Death Simulation Results:\n")
cat("===============================\n")
cat("Human outbreak probability:", round(human_outbreak$outbreak_probability, 3), "\n")
cat("Mean peak human infections:", round(human_outbreak$mean_peak, 0), "\n")
cat("Mean epidemic duration:", round(human_outbreak$mean_duration, 1), "years\n")

# Plot multi-host dynamics
plot_dynamics(results_black_death, compartments = c("I", "Ih", "Q"))
```

## Cross-Temporal Comparison

Compare historical vs modern plague dynamics:

```{r temporal-comparison, fig.cap="Historical vs modern plague transmission dynamics"}
# Compare different historical periods
historical_scenarios <- list(
  "Medieval (Black Death)" = "historical",
  "Early Modern" = "modern-estimates", 
  "Contemporary" = "defaults"
)

historical_results <- map(historical_scenarios, ~ run_plague_model(
  scenario = .x,
  n_particles = 50,
  years = 5
))

# Professional comparison using built-in function
plot_comparison(historical_results, compartment = "I")

# Quantitative comparison
historical_metrics <- historical_results |>
  imap(~ tibble(
    period = .y,
    outbreak_prob = .x |> 
      calculate_outbreak_metrics(compartment = "I") |> 
      summarize_outbreak_metrics() |> 
      pull(outbreak_probability),
    R0 = calculate_R0(load_scenario(historical_scenarios[[.y]]))
  )) |>
  bind_rows()

print(historical_metrics)
```

# Public Health Applications

## Control Intervention Analysis

Evaluate effectiveness of different plague control strategies:

```{r intervention-analysis, fig.cap="Quantitative assessment of control intervention effectiveness", eval = FALSE}
# Define intervention scenarios
interventions <- list(
  "Baseline" = list(),
  "Carcass Removal" = list(delta_R = 1.0),      # Faster carcass cleanup
  "Host Reduction" = list(K_r = 2000),           # Reduce rat populations
  "Reduced Contact" = list(beta_r = 0.3),        # Lower transmission rate
  "Combined Approach" = list(delta_R = 0.8, K_r = 2500, beta_r = 0.4)
)

# Run intervention scenarios
intervention_results <- imap(interventions, ~ {
  args <- c(.x, list(scenario = "modern-estimates", n_particles = 50, years = 6))
  do.call(run_plague_model, args)
})

# Calculate intervention effectiveness
intervention_metrics <- intervention_results |>
  imap(~ {
    metrics <- .x |> 
      calculate_outbreak_metrics(compartment = "I") |> 
      summarize_outbreak_metrics()
    tibble(
      intervention = .y,
      outbreak_probability = metrics$outbreak_probability[1],
      mean_peak = metrics$mean_peak[1],
      outbreak_reduction = 1 - (outbreak_probability / intervention_metrics$outbreak_probability[1])
    )
  }) |>
  bind_rows()

# Calculate effectiveness relative to baseline
baseline_prob <- intervention_metrics$outbreak_probability[1]
intervention_metrics <- intervention_metrics |>
  mutate(effectiveness = (baseline_prob - outbreak_probability) / baseline_prob * 100)

print(intervention_metrics[c("intervention", "outbreak_probability", "effectiveness")])

# Visualize intervention effectiveness
plot_comparison(intervention_results, compartment = "I")
```

# Summary

The `yersinia` package provides a comprehensive, professionally designed toolkit for plague transmission modeling that bridges theoretical epidemiology with practical public health applications.

## Key Capabilities Demonstrated

### **Modeling Framework**
- **Stochastic simulation**: Captures demographic noise and extinction/recolonization dynamics missing from deterministic models
- **Multi-host dynamics**: Carcass-based rat-to-human transmission with realistic biological parameters
- **Bayesian fitting**: Particle-filter pMCMC (`monty`) fits the same odin2 model to daily death-count time series

### **Professional Analysis Tools**
- **Core functions**: `calculate_outbreak_metrics()`, `calculate_R0()`, `calculate_force_of_infection()`, `calculate_equilibrium()`
- **Stable plotting**: `plot()` method, `plot_comparison()`, `plot_dynamics()`, `plot_phase_portrait()`, `plot_sensitivity()`
- **Parameter exploration**: Systematic sensitivity analysis and model comparison frameworks

### **Evidence-Based Parameters**
- **Curated scenarios**: Historical, contemporary, and research-validated parameter sets
- **Flexible customization**: Runtime parameter modification without file editing
- **Literature integration**: Direct implementation of parameters from key plague research

### **Real-World Applications**
- **Public health planning**: Control intervention assessment and effectiveness quantification
- **Historical modeling**: Black Death scenarios and cross-temporal epidemic comparison
- **Policy evaluation**: Evidence-based assessment of surveillance and control strategies

## Why Choose Stochastic Models?

Traditional deterministic models fail to capture:
- **Small population effects** where random events drive dynamics
- **Stochastic extinction and recolonization** driven by demographic noise
- **Uncertainty quantification** essential for policy decisions
- **Realistic outbreak variability** observed in natural systems

The `yersinia` package addresses these limitations while maintaining computational efficiency and analytical rigor.

## Getting Help & Further Resources

- **Function reference**: `help(package = "yersinia")`
- **Main modeling function**: `?run_plague_model`
- **Parameter details**: `print()` on loaded scenarios shows metadata and sources
- **Theoretical background**: `vignette("reference-deterministic-models")` for deterministic comparisons
- **Bayesian fitting walkthrough**: `vignette("monty")` for pMCMC fitting to historical death-count data
- **Demographic coupling**: `vignette("coupled-demography")` for the age-structured human + food-production + rat logistic model

## Key References

1. Didelot, X., Whittles, L. K., & Hall, I. (2017). Model-based analysis of an outbreak of bubonic plague in Cairo in 1801. *J. R. Soc. Interface*, 14: 20170160.
2. Keeling, M. J., & Gilligan, C. A. (2000). Metapopulation dynamics of bubonic plague. *Nature*, 407(6806), 903-906.
3. Stenseth, N. C., et al. (2008). Plague: past, present, and future. *PLoS Medicine*, 5(1), e3.

---

*The yersinia package emphasizes reproducible, evidence-based modeling with professional-grade analysis capabilities for epidemiological research and public health decision-making.*
